Activity of MukBEF for chromosome management in E. coli and its inhibition by MatP

Version 1

Abstract

Abstract

While different features for the activity of the bacterial canonical SMC complex, Smc-ScpAB, have been described in different bacteria, not much is known about the way chromosomes in enterobacteria interact with their SMC complex, MukBEF. Here we used a number of in vivo assays in E. coli to reveal how MukBEF controls chromosome conformation and how the MatP/matS system prevents MukBEF activity. Our results indicate that the loading of MukBEF occurs preferentially in newly replicated DNA, at multiple loci on the chromosome where it can promote long-range contacts in cis even though MukBEF can promote long-range contacts in the absence of replication. Using HiC and ChIP-seq analyses in strains with rearranged chromosomes, the prevention of MukBEF activity increases with the number of matS sites and this effect likely results from the unloading of MukBEF by MatP. Altogether, our results reveal how MukBEF operates to control chromosome folding and segregation in E. coli.

Introduction

SMC complexes play key roles in many processes involved in chromosome management, from genome maintenance, interphase chromatin organization, sister chromatids alignment, chromosome folding and condensing, to DNA recombination at specific stages of the cell cycle. (Yatskevich et al., 2019). A general model for SMC complex activity relies on their properties to bridge DNA elements and by doing so build DNA loops in cis and hold together the sister chromatids in trans; to do that, they bind DNA and processively extrude a DNA loop in a ATPase-driven loop extrusion, thereby compacting and organizing DNA (Davidson et al., 2019; Ganji et al., 2018; Kong et al., 2020). Yet many molecular features that determines the activity of the various SMC complexes are still unclear and it is not known whether they work using the same basal mechanisms (Bürmann et al., 2021; Davidson et al., 2019; Pradhan et al., 2022). For example, it is not yet clear whether I) the activity of SMC complexes is mediated by a single complex or if it involves cooperation between several complexes that organize into dimers or even oligomers (Badrinarayanan et al., 2012; Hassler et al., 2018), ii) the DNA extrusion involves a topological or nontopological mechanism, i.e. does the DNA pass through the SMC ring topologically (Bürmann et al., 2021; Davidson et al., 2019; Pradhan et al., 2022), iii) the activity of SMC complexes relies on the folding capacities of the SMC coiled-coil arms facilitating large-scale conformational changes (Bürmann et al., 2019), and iv) multiple complexes that encounter one another on the same DNA in living cells bypass each other or collide (Anchimiuk et al., 2021; Brandão et al., 2021). Cohesin and Condensin are the most characterized SMC complexes found in many eukaryotes. In bacteria, three different forms of SMC-like complexes defined as bacterial condensins have been identified, Smc-ScpAB, MukBEF, MksBEF; they are considered functionally related to condensins as they are thought to compact chromosomes and facilitate the segregation of sister chromosomes (Lioy et al., 2018, 2020; Marbouty et al., 2015). In bacteria, Smc-ScpAB represents the most highly conserved complex, while MukBEF and MksBEF represent diverged condensins (Cobbe and Heck, 2004; Yoshinaga and Inagaki, 2021).

Eukaryotic and prokaryotic SMC complexes are composed at least of 5 subunits: two Smc subunits, a kleisin subunit, and two additional subunits (kite and hawk subunits according to the type of SMC complex). Smc proteins associate with the kleisin protein to form a ring-shaped ATPase assembly. Additional subunits associate with this tripartite complex: the “Kite” family associates with bacterial and archaeal SMC complexes and also with the eukaryotic SMC5/6 complex while the “Hawk” family interacts with condensin and cohesin (Yatskevich et al., 2019). These additional subunits are thought to be required for the activity and to differentiate functions; for example, while condensin I and II share the same pair of Smc proteins, the difference in the subunit composition specifies their spatiotemporal dynamics and functional contributions to mitotic chromosome assembly (Hirota et al., 2004; Kong et al., 2020).

SMC function and dynamics on DNA requires additional auxiliary proteins (Baxter et al., 2019). More specifically, its loading on DNA and unloading of the DNA may depend on specific factors, at specific sites. For example, in Bacillus subtilis, the segregation ParB protein bound to parS site directly bind the Smc subunit; the ParB clamp presumably present DNA to the SMC complex to initiate DNA loop extrusion (Gruber and Errington, 2009; Sullivan et al., 2009; Wang et al., 2015, 2017). Upon translocation, the site-specific recombinase XerD bound to its binding site unloads SMC complexes in the terminus region of the chromosome and this process is thought to involve specific interactions between the different components (Karaboja et al., 2021).

MukBEF was the first SMC complex identified. In E. coli, MukBEF is thought to be required for chromosome segregation as muk mutants present many anucleate cells or mis-segregated chromosomes (Niki et al., 1991). How MukBEF may promote chromosome segregation and organization has remained elusive for a long time and is still not clear (Nolivos and Sherratt, 2014). The effect of the MukBEF complex in E. coli appears to be radically different from that of SMC in B. subtilis or other bacteria. Instead of aligning the chromosome arms from a centromere-like locus, MukBEF promotes DNA contacts in the megabase range within each replication arm (Lioy et al., 2018, 2020). MukBEF promotes long range interactions along the chromosome except in the Ter region where MatP prevents its activity (Lioy et al., 2018). Under conditions of increased chromosome occupancy of MukBEF, the E. coli chromosome appears to be organized as a series of loops around a thin (<130 nm) MukBEF axial core (Mäkelä and Sherratt, 2020). Whether MukBEF is loaded at a particular locus is still an open question. The complete atomic structure of MukBEF in complex with MatP and DNA has been determined by electron cryomicroscopy (Bürmann et al., 2021); it contains also the MukBEF binding partner AcpP protein (Prince et al., 2021). It revealed that the complex binds two distinct DNA double helices reminiscent of the arms of an extruded loop, MatP-bound DNA threads through the MukBEF ring, while the second DNA is clamped by MukF, MukE, and the MukB ATPase heads. The presence of MatP in the complex together with its ability to prevent MukBEF activity prompted authors to propose that MatP might be an unloader of MukBEF (Bürmann et al., 2021).

Here we have performed a number of experiments using different in vivo approaches to further characterize how MukBEF contributes to chromosome management in E. coli. We have used ChIP-seq and Chromosome Conformation Capture (3C-seq and HiC) experiments to study how MukBEF is loaded on the chromosome and promotes long range DNA contacts. By using strains with various chromosome configurations obtained by programmed genetic rearrangements, we have explored how MukBEF activity proceeds along the chromosome and how MatP bound to matS sites prevents its activity. Our results together with comparative genomics analyses allow us to address the biological significance of multiple matS sites in the Ter region of chromosomes in Enterobacteria and of the absence of MukBEF activity.

Results

MukBEF activity does not initiate at a single locus

In order to characterize how MukBEF interacts with the chromosome and initiate its activity of long-range contacts, it was necessary to set-up a system to reveal using HiC the appearance and spread of long-range contacts along the chromosome upon MukBEF synthesis. The rationale was based on previous findings showing that long-range DNA contacts within replication arms, outside the Ter region, result from MukBEF activity. If MukBEF loads at a specific locus as observed for Smc-ScpAB at parS sites, we would expect to detect the appearance and spreading of long-range contacts from this site upon MukBEF synthesis. By contrast, if MukBEF loads stochastically or at multiple loci on the chromosome, long-range contacts should occur at multiple sites.

The analysis of MukBEF activity required an efficient system to control its activity. To conditionally express MukBEF, the mukBEF operon was cloned onto a medium-copy number plasmid under control of a lacI promoter and introduced in a mukF mutant. In the absence of inducer, as observed for a mukF mutant, no growth was detected at 37°C and the amount of anucleated cells at 22°C was similar to that of the mukF mutant (13% versus 15%). Induced expression of mukBEF functionally complemented the absence of MukF, restoring growth at 37°C in LB (Figure 1A) and accurately segregating the chromosome at 22°C, as evidenced by the low amount of anucleate cells (less than 2%) observed in the presence of the inducer (Figure 1B).

MukBEF activity detected along the chromosome

(A) Complementation of the ΔmukF mutant. Serial dilutions (10-1 to 10-5) of an exponential culture of MG1655 (WT), MG1655 ΔmukF ppSV38, and MG1655 ΔmukF ppSV38::mukBEF were plated on media with (Right) or without (Left) inducer (IPTG). Plates were incubated at a permissive temperature of 22°C (top) or at a non-permissive temperature of 37°C (bottom).

(B) : Percentage of anucleate cells (blue bars) in wild type (WT) and mukF mutant strains, complemented or not by the plasmid ppsV38::mukBEF, grown in minimal medium at 22°C with or without IPTG. Anucleate cells were identified and counted after DAPI staining. The histograms and error bars represent the means and standard deviations from at least three independent experiments.

(C) : Normalized Hi-C contact maps (5Kb bin resolution) obtained from a ΔmukF strain complemented with ppsV38::mukBEF after different induction times (0 min, 40 min, 2 h). Cells were grown in permissive conditions at 22°C in minimal medium. The X and Y axes represent genomic coordinates in megabases (Mb).

(D) Ratio of normalized contact maps of ΔmukF ppSV38::mukBEF grown in the presence and absence of ITPG, represented in the left panel for 40 minutes of growth. The right panel shows the ratio of normalized contact maps for 2 hours of growth with and without ITPG. A decrease or increase in contacts in the induced condition compared with the non-induced condition is represented with a blue or red color, respectively. The black line represents a schematic chromosome, with the Ter domain highlighted in light blue.

(E) Quantification of the range of cis contacts of chromosomal loci along the chromosome of a ΔmukF ppsV38::mukBEF strain grown at 22°C under four different conditions: light blue (without ITPG), red (after 20 minutes of ITPG), yellow (after 40 minutes of ITPG), and purple (after 2 hours of ITPG).

To determine how MukBEF activity initiate in the E. coli chromosome, we induced the expression of mukBEF and monitored the appearance of long-range DNA contacts at different times after induction. HiC contact maps were established 20 min, 40 min and 2 hours after induction and compared to that obtained in the absence of induction (Figure 1C – figure supplement 1A, B): long-range DNA contacts were hardly visible after 20 min (Figure 1C -figure supplement 1A), were readily observed after 40 min and reached after 2 hours a level similar to that observed in wild-type strains (figure supplement 1C). The ratio of normalized contact maps of the induced strain at different time points to the non-induced strain allowed to visualize the presence of long-range contacts all over the chromosome except in the Ter macrodomain (Figure 1D).

The range of DNA contacts along the chromosome was quantified by measuring the width of the diagonal perpendicular to it, using an adapted version of the quantification method developed by (Wang et al., 2017 ; Lioy et al., 2020). The plot obtained (Figure 1E) allows to estimate the effect of MukBEF activity on all loci of the E. coli genome. This confirmed the increase in the range of contacts following induction of MukBEF predicted from the ratio of normalized contact maps, with 40 minutes necessary to observe a significant effect and 2 hours to reach the maximum range of contact, as observed in the wild-type strain (Lioy et al., 2020) (Figure supplement 1). As expected, the Ter region was not affected by the induction of MukBEF because of the presence of MatP bound to matS sites.

Altogether, the results indicate that the increase in contact does not originate from a specific position on the chromosome but rather appears from numerous sites, suggesting that the activity of MukBEF does not initiate at a single locus but rather from multiple loci along the chromosome.

MukBEF activity initiates in different regions of the E. coli genome

The MatP/matS system has been shown to prevent MukBEF activity in the Ter MD. We took advantage of this property to unravel mechanistic aspects of MukBEF activity and address the following issues: could MukBEF interacts with a region flanked by different Ter segments and could MukBEF operate from different regions of the chromosome. To answer these questions, we used bacteriophage λ site-specific recombination as described before (Thiel et al., 2012) to perform large transpositions of a segment of the right (“RiTer”) and left (“LiTer”) replichore at different loci (figure supplement 2A), in the Ter domain, and analyze MukBEF activity in the resulting strain (Figure 2A).

MukBEF activity initiates at different regions of the E.coli chromosome

(A) (left) Normalized 3C-seq 5kb bin contact map of strain RiTer and (right) normalized Hi-C 5Kb bin contact map of strain LiTer15. The positions of the two Ter segments are highlighted in blue on the scheme, and the dashed lines project their positions onto the matrix.

(B) Ratio of normalized contact maps of RiTer strain on RiTer ΔmukB strain (left), or RiTer on RiTer ΔmatP (middle), and LiTer15 on LiTer15 ΔmukB strain(right). A decrease or increase in contacts in the transposed cells compared to the transposed mutant cells is shown in blue or red color, respectively.

(C) The graph shows the quantification of the Hi-C diagonal width for loci along the chromosome of the RiTer strain (left panel, blue line) and the RiTer ΔmukB strain (left panel, red line), as well as LiTer15 (right panel, blue line) and LiTer15 ΔmukB strain (right panel, red line). Schematic representations of the two Ter segments are indicated below the graph, highlighting the fact that the range of contacts decreased on all Ter fragments.

The transposition in the RiTer configuration results in a region of 450kb devoid of matS, located between two Ter segments of 378 Kb and 437kb, which contain 11 matS and 17 matS respectively (figure supplement2A). 3C-seq experiments were performed with this strain (“RiTer”) and on the same strain deleted for matP or for mukB (Figure 2A-suplementary Figure 2B). The ratio of normalized contact maps of the RiTer strain to the RiTer ΔmatP allowed to visualize the reduction of long-range contacts in the two Ter segments in the presence of MatP (Figure 2B). On the other hand, the ratio of normalized contact maps of the RiTer strain to the RiTer ΔmukB allowed to visualize the increase of long-range contacts between and outside the two Ter segments in the presence of MukBEF. Remarkably, long-range DNA contacts specific of MukBEF activity were observed over the chromosome except in the two Ter segments. The range of DNA contacts along the chromosome were quantified by measuring the width of the diagonal in the two strains; these plots (Figure 2C) allow to estimate the effect of MukBEF activity on all loci inside and outside Ter segments. These results indicated that MukBEF can operate to form long-range contacts from a segment of a Right MD between the two Ter segments and that MatP can prevent MukBEF activity when segments carrying matS sites are moved in different locations of the genome.

Similar experiments were performed with a strain carrying a transposed segment from the Left replichore in the Ter MD (LiTer 15). In this configuration, a region of 1085 kb devoid of matS, is flanked by two Ter segments of 427 Kb and 374 kb containing 13 matS and 15 matS respectively (figure supplement 2A). HiC experiments were performed on this strain (“LiTer”) (Figure 2A). As observed with the RiTer strain, long-range DNA contacts specific of MukBEF activity were observed over the chromosome except in the two Ter segments. The ratio of normalized contact maps of the LiTer15 strain to the LiTer15 ΔmukB allowed to visualize the increase of long-range contacts outside the two Ter segments in the presence of MukBEF (Figure 2B and 2C-figure supplement 2C).

Altogether, these results demonstrate that MukBEF activity can be initiated from multiple regions of the chromosome, in the Right, Left or Ori domains, and that Ter segments cannot insulate DNA segments devoid of matS sites. They also reveal that MukBEF does not translocate from the Ori region to the terminus of the chromosome as observed with Smc-ScpAB in different bacteria. Furthermore, the results indicated that Ter segments carrying 17 and 11 matS can prevent MukBEF activity and restrict DNA contacts.

matS determinants to prevent MukBEF activity

The method described above, splitting the Ter region in two parts, revealed that two parts of Ter can prevent MukBEF activity. Thus, by varying the way Ter is spilt in two parts, one should be able to identify matS determinants required to affect MukBEF activity.

Two previous studies differ slightly in the prediction for the number and consensus sequence of matS (Mercier et al., 2008; Nolivos et al., 2016).To clarify this, three independent ChIP-seq experiments were performed, revealing 28 matS and a newly derived consensus sequence (see figure supplement 3). To examine the requirements for the inhibition of MukBEF by MatP, a 1 Mb region of the left replichore was transposed at different positions into the TER domain, thus dividing Ter into two parts (referred to as Ter1 and Ter2, Ter1 comprising the dif site and the terminus of replication) of different sizes and carrying different number of matS (Figure 3A). Three transpositions were performed generating Ter2 domains of 253, 209 and 136 kb containing 9, 7 or 4 matS, respectively (strains LiTer9, LiTer7, LiTer4). We used Hi-C to test the ability of these Ter2 regions to inhibit MukBEF (LiTer9 in figure supplement 4A, LiTer7 and LiTer4 in Figure 3B). Long-range DNA interactions were readily observed on both regions flanking Ter2 in both three strains, and long-range contacts are affected in the Ter2 segment, even when only 4 matS were present (Figure 3B). Similar HiC experiments were performed with the same strains deleted for matP (figure supplement 4B). The ratio of normalized contact maps of the LiTer4 strains to the corresponding matP derivatives (Figure 3C) clearly revealed that long-range contacts were limited in Ter 2 of the different strains. Remarkably, the 136 kb region carrying 4 matS sites in LiTer4 was sufficient to decrease long-range contacts promoted by MukBEF activity.

The number and distribution of matS sites have different effects on MukBEF Inhibition

(A) The figure shows a schematic representation of different transpositions with the number of matS sites located on each Ter region. The three att sites are integrated into the chromosome in the same orientation, and attL/attR are fixed positions on all LiTer transpositions. attB is inserted at different positions into the Ter allowing, upon transposition, the division of the Ter domain into two subdomains, Ter1 and Ter2, containing different numbers of matS sites. The number of matS sites is indicated for the different transpositions, and the distribution of matS on the Ter2 segment is indicated in the schematic for the three transposed strains, LiTer4, LiTer7, and LiTer9.

(B) Normalized Hi-C contact map with 5Kb bin resolution of the transposed strains LiTer7, LiTer4, and LiTer4 ΔmatS28. The position of the different Ter regions is highlighted below the matrix, and by dashed lines on the matrix.

(C) Ratio of normalized 5Kb bin contact maps for the different transpositions compared to the matP mutant on the same genetic organization. The position of the different Ter regions is highlighted below the ratio, and by dashed lines on the ratio.

(D) Quantification of the Hi-C diagonal width for loci along the chromosome for the transposed strains LiTer7, LiTer4, the derivative mutant LiTer4ΔmatS28, and LiTer4ΔmatP. The schematic map below represents the LiTer7 configuration. The position of the different Ter regions is highlighted below the graph, and by dashed lines on the graph.

(E) This panel quantifies the range of contacts in the Ter2 region or in the corresponding sequence on the wild-type (WT) configuration. Boxplot representations are used, indicating the median (horizontal bar), the 25th and 75th percentiles (open box), and the rest of the population.

The ability of different Ter segments to inhibit the formation of long-range DNA contacts was quantified by measuring the range of contacts in the different Ter segments (Figure 3D and E). The median range of contact for the Ter segment matS20-matS28 in the WT condition is 263 +/−24kb. Remarkably, no significant differences in the range of contacts (275 +/−9kb) were observed in strain Liter9 for the Ter2 segment carrying 9 matS; it is noteworthy that this region carrying 9 matS sites shows the same range of contacts (280 +/−35kb) in the wt configuration. This result shows that an transposed 253-kb region carrying 9 matS prevent MukBEF activity as much as the same segment present in the 838-kb long Ter MD.

To further characterize how MatP/matS can prevent MukBEF activity, we measured the range of contacts in Ter2 segments carrying 7 or 4 matS sites. For the 209-kb segment carrying 7 matS sites, the range of contact was increased to 330 +/−9kb while it increased to 352 +/−12kb for the 136-kb segment with 4 matS. Values obtained for these segments when present as part of the WT Ter MD were slightly lower (Figure 3E). These results indicate that segments carrying 7 or 4 matS affect MukBEF activity, even though at a lower level than a 250-kb segment with 9 matS sites.

The four matS sites present in the 136-kb Ter2 region of LiTer4 are not distributed regularly in that region (Figure 3A). We took advantage of this irregular spacing to test the capacity of a segment carrying 3 matS sites to prevent MukBEF activity and to explore whether the density of matS sites can modulate MukBEF activity. To address these two questions, we deleted either matS26 or matS28from strain LiTer4 and probed using HiC the long-range of contacts. In strain LiTer4 ΔmatS26, three matS sites are distant from 78 and 58 kb whereas the matS sites in LiTer4 ΔmatS28strain are separated by 13 and 65 Kb. The ratio of normalized contact maps of the LiTer4 ΔmatS26 or LiTer4 ΔmatS28 strains to the corresponding matP derivative (Figure 3B, 3C and figure supplement 4C) revealed that long-range contacts were affected by the presence of only 3 matS sites in the 136-kb segment. By measuring the range of contacts in strains LiTer4 ΔmatS26 and LiTer4 ΔmatS28 (Figure 3E), the density of three matS sites provided by matS25-matS26 and matS27 in strain LiTer4 ΔmatS28 seems to be as efficient (range contact of 357 kb +/−11 kb) as the four matS sites in strain LiTer4 (range contact of 352 +/−12kb). By contrast, in LiTer4 ΔmatS28, the range of contact is increased to 370 kb +/−6 kb indicating a reduced ability to prevent MukBEF activity. Altogether, these results suggested that the density of matS sites in a small chromosomal region has a greater impact than dispersion of the same number of matS sites over a larger segment.

MukBEF preferentially binds in newly replicated regions

Results presented above indicated that MukBEF activity can be initiated from multiple regions of the chromosome and that, unlike Smc-ScpAB, MukBEF does not initiate its activity in the Ori region and translocate linearly to the terminus of the chromosome. To further characterize the loading and translocation process of MukBEF, we performed ChIP-seq experiments using a FLAG version of MukB on synchronized cells using a dnaC2 thermosensitive allele that allows to control the timing of replication initiation (Figure 4A). The cells were grown at a permissive temperature, then shifted to 40°C for 2 hours to allow the ongoing round of replication to complete without being able to initiate a new round; the cells were then shifted back to 30°C and samples were taken for ChIP-seq analyses after 10, 20 and 40 minutes. At t0, there is no variation in the number of sequences along the chromosome indicating the absence of replication. In this condition, the ChIP-seq signals show a slight enrichment of signals outside the Ter region. After 10 minutes at the permissive temperature, a fraction of >500 kb centered on oriC was replicated as revealed by an increase of sequencing reads in this region. After 20 minutes, a large zone of over 1.4 Mb was replicated, and after 40 minutes, the chromosome was fully replicated and the replication profile was similar to that of non-synchronized cells because of multiple new replication cycles initiated. Normalized ChIP-seq experiments were performed by normalizing the quantity of immuno-precipitated fragments to the input of MukB-Flag and then divide by the normalized ChIP signals at t0 to measure the enrichment trigger by replication. This experiment showed a 2- to 4-fold enrichment in the regions that have been replicated (Figure 4A). After 10 min, the signal was increased over 500 kb on each side of oriC. At 20 min, the signal progressed and corresponded to the regions that have been replicated. At 40 min, when the chromosome has been fully replicated, the signals obtained in the ChIP-seq samples indicated an enrichment of MukBEF all along the chromosome except in a 1.5 Mb region centered on the Ter region. As shown in Figure 4B, MukBEF enrichment drops to background levels 250kb before the Ter (Figure 4A). This enrichment followed the progression of replication and spread from oriC towards the Ter. Altogether, these results suggest that MukBEF is loaded in newly replicated regions, progressively behind the replication process, except in the Ter region from which it would be excluded.

MukBEF preferentially binds in newly replicated regions and is exclude from Ter sequence

(A) Immunoblot analysis of mukB-flag dnaC2 strain. Exponentially growing cells were synchronized by incubating them at 40°C for 2 hours, followed by a shift to 30°C for 0, 10, 20, or 40 minutes. The replication progression was monitored by plotting the input sequence (left panel) in 50 kb bins. Normalized ChIP (ChIP over input) values for 50 kb bins are presented in the right panel, with the red asterisk indicating the peak observed in all E. coli ChIP-seq experiments. The positions of oriC and Ter are highlighted with a red dashed line and on the chromosome schematic below the figure.

(B) Immunoblot analysis was conducted on the LiTer7 mukB-flag dnaC2 strain. We observed that replication restart in this transposed strain exhibited a 20-minute lag. Therefore, cells in replication stop state were shifted to 30°C for 60 minutes to achieve a comparable replication progression to the 40-minute WT strain, as shown by the plotting of the input sequence (left panel). MukB enrichment generally followed replication progression, except in the two Ter regions, as demonstrated by the immunoblot of MukB ChIP normalized by input at 50 kb bins (right panel).

(F) Normalized Hi-C 5 kb bin contact map of ΔmukF ppsV38::mukBEF dnaC2 strain. Exponential cells were incubated at 40°C for 2 hours to prevent replication initiation (non-replicating condition) and were then shifted to 30°C for 40 minutes (right panel). MukBEF induction was performed by adding IPTG to the media for 2 hours.

Ter segments prevent MukBEF binding

To further explore the ability of MatP to exclude MukBEF from the Ter region, we tested the ability of MatP to exclude MukBEF from chromosomal regions containing matS sequences. Chip-Seq experiments were performed on the Liter7 strain, which has a Ter2 segment closer to oriC. As in the wild-type background, MukBEF is preferentially associated with newly replicated sequences and shows a 3-fold increase following replication fork progression. However, a break in this enrichment profile is detected in the sequence corresponding to Ter2 (Figure 4B). These results suggest that MukBEF does not bind or persist in segments carrying matS sites, and that the MatP/matS system prevent residence of MukBEF in that region.

Long-range contacts correlate with MukBEF binding

Since MukBEF was shown to bind preferentially in newly replicated regions, we wanted to test if a preferential activity of MukBEF was detectable in those regions. To do this, we performed Hi-C on non-replicating cells lacking MukF, and induced MukBEF expression with or without restarting replication. In the absence of replication and of MukBEF, long-range contacts were constrained by barriers that delimit domains, called Chromosome-Interacting Domains (CIDs), previously detected in WT cells in growing conditions (Lioy et al., 2018; figure supplement 5). Remarkably, after two hours of MukBEF induction in non-replicating cells, long-range contacts were detected except in the Ter region, indicating that MukBEF activity does not require newly replicated DNA to promote long-range contacts (Figure 4C). Finally, restarting replication for 40 minutes after 2 hours of MukBEF induction did not alter significantly the range and distribution of long-range DNA contacts observed in the absence of replication but in the presence of MukBEF. Altogether, these results indicate that MukBEF promotes long-range DNA contacts independently of the replication process even though it binds preferentially in newly replicated regions.

Functional implications, comparative genomics of matS distribution

In E. coli, there are 28 matS sequences dispersed throughout the 1.03 MB Ter domain. The average matS density is 3 matS per 100 Kb, but this distribution is not uniform and the matS density doubles in the vicinity of the dif sequence (figure supplement 6A). Despite the fact that only 9 matS are sufficient to completely inhibit MukBEF activity, matS sites have been selected and consequently a large portion of the E. coli chromosome is inaccessible to MukBEF.

To determine if other species also possess a large Ter domain, we used the matS consensus sequence to identify the Ter domain in 16 bacteria from the enterobacterals, pasteurellales, and vibrionales groups with the higher number of sequenced genomes. The Ter domain was defined as the longest stretch of matS sequences between areas that are at least 100 KB devoid of matS. The size of the Ter domain varies from 300 KB to 1 MB, representing 6 to 25% of the chromosome, and contains 6 to 77 matS (Table 1). In most of these species, more than 14 matS sequences are distributed over 540 KB, and the Ter domain always contains the dif sequence. To test if the matS distribution might differ in the vicinity of the dif sequence, we measured the matS density and centered the distribution on the dif sequence. As shown in Figure 5, the number of matS per kb increases for all species and reaches its maximum near the dif site.

Distribution of matS sites culminates at dif across γ-proteobacteria

The density of matS sequences over a 100kb-unit region was measured in 16 γ-proteobacteria, and the resulting distribution is plotted in the figure. The red line represents the 75th percentile of this distribution, while the blue line represents the 25th percentile. The light green area between these two lines represents the 50% of values closest to the median. The distribution is centered on the dif site, and the x-axis represents the genomic distance in megabases (Mb) from dif, while the y-axis represents the number of matS sequences per 100 kilobases (Kb).

Discussion

SMC complexes play a fundamental role in the organization of genomes in all domains of life. Different models involving their loading, translocation or extrusion of DNA loops as well as their unloading have been proposed, based on results obtained with different complexes and in different models. Among the SMC complexes, MukBEF has specific features: MukBEF is only found in enterobacteria and some related bacterial genera where it is involved in chromosome segregation; the mukBEF genes belong to a group of genes co-occurring with the Dam methylase gene including also matP and other genes involved in DNA metabolism; MukBEF exists as dimers of dimers connected by the kleisin subunit MukF; by its activity, MukBEF does not align the two arms of the chromosome like the canonical bacterial Smc-ScpAB complex but instead promotes long distance contacts in cis like a eukaryotic condensin. Remarkably, MukBEF activity is not detected within the specific chromosomal Ter domain, one-fifth of the E. coli chromosome, due to the presence of MatP associated with this region. Although details about MukBEF and its activity have been revealed in recent years, key steps in its function remain to be characterized, including loading onto the DNA molecule, its actual loop extrusion activity, and its unloading by MatP.

Altogether, our data provide an integrated view of MukBEF activity to organize the E. coli chromosome. By different ChIP-seq and HiC approaches performed in different strains, some of which have undergone programmed genetic rearrangements, we showed that MukBEF loading does not only involve the Ori region but also different regions of the chromosome. Our results also indicated that although MukBEF binds preferentially in newly replicated regions, its activity is detected even in the absence of replication and long-range contacts appear similarly throughout the genome, except in the Ter region. These results support a model in which MukBEF molecules are bound to the chromosome, molecules are removed or displaced by replication, MukBEF molecules readily reassociate in newly replicated regions except in Ter region in which the unloading of MukBEF is enhanced by MatP bound to matS sites. These results unveiled that fluorescent MukBEF foci previously observed associated with the Ori region were probably not bound to DNA (Nicolas et al., 2014). Instead, our results further support the previous proposal that MukBEF may organize a series of loops around a thin MukBEF axial core. Altogether, our results reveal a striking contrast with the way Smc-ScpAB loads on DNA by interacting with the ParB at parS sites and then translocates towards the ter region. Instead, long-distance contacts promoted by MukBEF did not occur as a wave from a specific region but rather initiate at different positions, in different regions of the genome. Thus, the activity of MukBEF appears to be more similar to that of eukaryotic condensins than to that of the Smc-ScpAB complex.

Our results showed that MukBEF was not detected in regions containing matS sites and that these regions were devoid of contacts extending over 600 kb. By varying the number of matS sites at an ectopic position, we showed that MukBEF inhibition is graded with the number of matS; while an effect is already visible with 3 matS scattered over a 78 kb region, the maximal effect seems to be reached with 9 matS sites distributed over a 253 kb region. As proposed for Smc-ScpAB and its unloading XDS site, if the movement of DNA in the loop extrusion process involves large steps, it is conceivable that several matS sites are required for a complex to be trapped and discharged by MatP bound to a matS site. Accordingly, the density of matS sites seems to affect the inhibition efficiency supporting this assumption. Further experiments will be needed to analyze in details the optimal spacing of matS sites to inhibit MukBEF activity.

The structure of the MukBEF-AcpP-MatP/matS complex obtained by cryoEM revealed the entrapment of two topologically separated DNA segments in two distinct compartments called “ring” and “clamp”, with the matS site present in the ring compartment. The proposed model for MukBEF unloading stipulates that the unloading of the segment carrying matS site in the ring compartment is coupled to the unloading of the other segment in the clamp. One of the MatP monomers forms a contact with one of the MukE monomers while the joint binds and positions MatP between the MukB arms. The joint interface is much larger than the MukE-MatP bridge and likely provides the major binding energy for association (Bürmann et al., 2021). Indeed, a change to alanine of the five residues between H38 and D42 in contact with MukE did not affect MatP’s ability to inhibit MukBEF activity (figure supplement 7). Refined experiments will be required to assess the outcome of mutating MatP residues involved in the interaction with the joint as mutations of those residues also affect matS binding (Dupaigne et al., 2012).

MukBEF activity is detected by the appearance of contacts at a distance close to the megabase and such contacts appear on about 4 Mb. Even if the molecular details of this activity remain to be characterized, it is tempting to speculate that MukBEF molecules can be discharged from DNA in the absence of MatP, i.e. outside the Ter region. Furthermore, it can be noted that MukBEF activity does not appear to be significantly disrupted in a matP mutant suggesting that MukBEF can be discharged from DNA in the absence of MatP. As proposed by Bürmann and colleagues (Bürmann et al., 2021), MatP would act as a structural element that ensures ideal positioning of DNA close to the exit gate in the MukBEF complex that might ensure an efficient unloading of MukBEF from the DNA. Altogether, the results would indicate that MatP is not the MukBEF unloader per se, but rather that its ability to prevent MukBEF activity has been selected to protect the Ter region from a condensin activity.

The distribution of matS sites over a large region of the chromosome in enterobacteria results in the inhibition of MukBEF in Ter. MatP has already been shown to confer another property to Ter, through an interaction of its C-terminus with the septum-associated protein ZapB and localizing Ter at midcell. Remarkably, these two properties are independent as inhibiting MukBEF does not require its anchoring of Ter at the septum of division (Lioy et al., 2018). Given that the number of matS sites far exceeds the number required to prevent MukBEF activity and that the density of matS sites increases as we approach dif, we may speculate that MukBEF presence is mostly banished from the dif region. Two activities of DNA metabolism occur at this locus: resolution of chromosome dimers by XerC-XerD recombinases and the post-replicative decatenation of circular chromosomes. A major challenge for the future is to define whether and how MukBEF may interfere with one or both of these processes.

In the absence of replication and of condensin, the contact map displayed a single strong diagonal composed of very well-defined CIDs (for Chromosomal Interacting Domains), ranging in size from 20 to 400 kb. The CIDs clearly visible in populations of cells growing exponentially were prominent in these conditions revealing the impact of transcription on the structuring of the genome. Upon replication, the CID organization is less apparent because of the new contacts that occur between loci belonging to different CIDs. The induction of MukBEF in the presence of replication further attenuates the patterning of CIDs. Altogether, these results highlight the respective contribution of these three processes, transcription, replication and condensin activity, on the organization of bacterial genomes.

Materials and methods

Bacterial strains and plasmids

All E. coli strains used in this study are derived from E. coli MG1655. Deletion mutants were constructed as described in (Datsenko and Wanner, 2000). Mutations were combined via P1 transduction. A plasmid capable of synthesizing MukBEF was constructed by cloning the entire mukBEF operon into the ppSV38 plasmid using EcoRI/XbaI.

Media and growth conditions

(A) E. coli cells were cultured at 22°C and 30°C in either Lennox Broth (LB) or liquid minimal medium A (MM) supplemented with 0.12% casamino acids and 0.4% glucose. When necessary, antibiotics were added to the growth media at the following concentrations: ampicillin (Amp) at 100 μg/mL, kanamycin (Kan) at 50 μg/mL, chloramphenicol (Cm) at 15 μg/mL, spectinomycin (Sp) at 50 μg/mL, apramycin (Apra) at 50 μg/mL, and zeocin (Zeo) at 25 μg/mL.

3C-seq protocol

3C-seq librairies were generated as described (Lioy et al., 2018). Briefly 100ml of culture were crosslinked with formaldehyde (7% final concentration) for 30 min at room temperature (RT) followed by 30 min at 4°C. Formaldehyde was quenched with a final concentration of 0.25 M glycine for 20 min at 4°C. Fixed cells were collected by centrifugation and stored at −80°C until use. Frozen pellets consisting of approximately 1-2 x 109 cells were thawed and suspended in 600 μl of TE buffer (10 mM Tris-HCl, 0.5 mM EDTA, pH 8) with 4 μl of lyzozyme (35 U/μl). The mixture was incubated at room temperature for 20 minutes. Subsequently, SDS was added to a final concentration of 0.5% and the cells were incubated for an additional 10 minutes at room temperature. Lysed cells were then diluted 10 time in several tubes containing 450 μl of digestion mix (1X NEB 1 buffer, 1% Triton X-100). 100 units of HpaII were added and the tubes were incubated for 2 hours at 37°C. To stop the digestion reaction, the mixture was centrifuged for 20 minutes at 20,000 g, and the resulting pellets were resuspended in 500 μl of sterile water. The resulting digested DNA (4 ml in total) was divided into four aliquots and diluted in 8 ml of ligation buffer (1X ligation buffer NEB without ATP, 1 mM ATP, 0.1 mg/mL BSA, 125 units of T4 DNA ligase, 5 U/μl). Ligation was performed at 16°C for 4 hours, followed by overnight incubation at 65°C with 100 μl of proteinase K (20 mg/ml) and 100 μl EDTA 500 mM. DNA was then precipitated with an equal volume of 3 M Na-Acetate (pH 5.2) and two volumes of iso-propanol. After 1 hour at −80°C, the DNA was pelleted and suspended in 500 μl of 1X TE buffer. The tubes were incubated directly with 50 μl of proteinase K (20 mg/mL) for an overnight period at 65°C. Subsequently, all tubes were transferred to 2 ml centrifuge tubes and extracted with 400 μl of phenol-chloroform pH 8.0. The DNA was then precipitated, washed with 1 ml of 70% cold ethanol, and resuspend in 30 μl of 1X TE buffer in the presence of RNase A (1 μg/ml). The tubes containing ligated DNA (3C libraries), digested DNA, and non-digested DNA were pooled into three separate tubes. The efficiency of the 3C preparation was evaluated by running a 1% agarose gel.

Hi-C protocol

Hi-C librairies were generated as described (Thierry and Cockram, 2022).

108 cells growing in the exponential growth phase of E. coli were chemically crosslinked by the addition of formaldehyde directly to the cultures (3% final concentration) for 30 min at room temperature with gentle agitation. Crosslinking was quenched by the addition of glycine (0.5 M final concentration) for 20 min at room temperature. Celle were wash in 50 ml PBS 1X and centrifuged. The pellet was resuspended in 1 ml 1 x PBS and transferred to a 1.5 ml Eppendorf tube before a final centrifugation step (4000 x g, 5 min, room temperature), the supernatant was then removed and the pellet stored at −80°C. The pellet was then resuspended in 1 ml of 1x TE + complete protease inhibitor cocktail (EDTA-free, Sigma Aldrich) and transferred to a 2 ml Eppendorf tube with 4 µl of Ready to lyse Lysozyme for 20mn at room temperature. SDS was added to a final concentration of 0.5% and the cells were incubated for an additional 10 minutes. In a 10 ml Falcon tube, DNA was then prepared for digestion by the addition of 3 ml H2O, 500 μl 10X Digestion buffer (200 mM Tris-HCl pH 7.5, 100 mM MgCl2, 10 mM DTT, 1 mg/ml BSA) and 500 μl 10% Triton X-100 (Thermo-Fisher). After thoroughly mixing the reaction, 400 μl were removed and transferred to a 1.5 ml Eppendorf tube as a non-digested (ND) control. The restriction enzyme HpaII (New England Biolabs, 1000 U) was then added to the remaining sample and the tube incubated with gentle agitation for 3h at 37°C. The solution was centrifuged (16,000 x g, 20 min, room temperature. After removing the supernatant, the pellet was resuspended in 400 μl H2O, completed subsequently by adding 50 μl 10x Ligation Buffer (500 mM Tris-HCl pH 7.5, 100 mM MgCl2, 100 mM DTT), 4.5 μl 10 mM dAGTTp, 37.5 μl Biotin-14-dCTP (Thermo Fisher), 50 Units of DNA Polymerase I - Large Klenow Fragment (New England Biolabs). After briefly mixing, the reaction was incubated with agitation for 45 min at 37°C. The ligation was set up by adding the following; 120μl 10x Ligation Buffer, 12 μl 10 mg/ml BSA, 12 μl 100mM ATP, 540 μl H2O, 480 U T4 DNA ligase (Thermo Fisher). The reaction was mixed gently and then incubated with gentle agitation for 3h at room temperature. Following ligation, proteins were denatured by the addition of 20 μl 500 mM EDTA, 20 μl 10% SDS, and 100 μl 20 mg/ml proteinase K (EuroBio). The following day, DNA was purified using the standard procedure describe for the 3C-seq.

Chip-seq protocol

109 cells in the exponential growth phase of E. coli were chemically crosslinked by adding formaldehyde directly to the cultures (final concentration of 1%) for 30 minutes at room temperature with gentle agitation. Crosslinking was quenched by adding glycine (final concentration of 0.25 M) for 15 minutes at room temperature. The cells were washed twice with 1X TBS (50mM Tris-Hcl ph 7.6, 0.15M NaCl) and the pellet was resuspended in 1 mL of 1X TBS before a final centrifugation step (4000 x g, 5 minutes, room temperature). The supernatant was then removed and the pellet was stored at −80°C.

The pellet was resuspended in 500 µl of lysis buffer 1 (20% sucrose, 10 mM Tris pH 8, 50 mM NaCl, 10 mM EDTA) and 4 µl of ready-to-lyse lysozyme was added, followed by incubation at 37°C for 30 minutes. Then, 500 µl of lysis buffer 2 (50 mM Tris-HCl pH 4.7, 150 mM NaCl, 1 mM EDTA, 1% Triton X100) and a tablet of antiprotease cocktail (Roche) were added. The solution was transferred to a 1 mL Covaris tube and sonicated for 10 minutes (peak incident power 140 W/duty cycle 5%/cycle per burst 200). Cell debris were eliminated by centrifugation at 13,000 rpm for 30 minutes at 4°C. The supernatant was transferred to a fresh Eppendorf tube and 50 µl was used as input. The rest of the cell extract was mixed with 40 µl of anti-FLAG M2 resin previously washed in TBS and resuspended in lysis buffer 2. The solution was incubated overnight at 4°C with gentle mixing and then centrifuged for 30 seconds at 5,000 g. The pellet was washed twice with TBS containing 0.05% Tween 20 and three times with TBS.

To elute the immunoprecipitation, 100 µl of 1X TBS containing 15 µg of 3X FLAG peptide was mixed with the resin and incubated for 30 minutes at 4°C. Then, the solution was centrifuged for 30 seconds at 5,000 g and the supernatant was transferred to a new tube. A second elution step was performed on the resin before decrosslinking. IP and input was purified using MinElute Qiagen columns and then sequenced.

Processing of libraries for Illumina sequencing

The samples were sonicated using a Covaris S220 instrument to obtain fragments ranging in size from 300 to 500 base pairs. These fragments were then purified using AMPure XP beads and resuspended in 10 mM Tris-HCl. For Hi-C libraries, a biotinylated pull-down step was performed by adding 30 µl of streptavidin C1 MyOne Dynabeads from Invitrogen to 300 µl of binding buffer (10 mM Tris-HCl, pH 7.5, 1 mM EDTA, 2 M NaCl) and mixing with the Hi-C libraries for 15 minutes. DNA ends were then prepared for adaptor ligation following standard protocols as described in (Thierry and Cockram, 2022).

The Illumina sequencing process was performed in accordance with the manufacturer’s recommendations, with 15 cycles of amplification. The size of the DNA fragments in the libraries was assessed using TAE 1% agarose gel and tape station, followed by paired-end sequencing on an Illumina sequencer.

Chip-seq analysis

Between 10 to 20 million reads were recovered for each sample. We used Bowtie2 software to perform mapping in local mode, and the mpileup software from Samtools to calculate coverage for each position in the genome. Normalized ChIP was then calculated by normalizing the number of reads at each position by the total number of reads, and dividing this number by the normalized input. A sliding window of 50 kb was applied to smooth the variations. Peaks for MatP ChIP-seq were identified by extracting positions where the number of reads was ten times higher than the background for at least 30 consecutive base pairs. The center of these distributions was considered the center of the peak. The sequences were then extracted and used on the MEME suite to identify a common motif.

Generation of contact maps

Contact maps were constructed as described previously (Lioy et al., 2018). Briefly, each read was assigned to a restriction fragment. Non-informative events such as self-circularized restriction fragments, or uncut co-linear restriction fragments were discarded, as in (Cournac et al., 2012). The genome was then divided into 5 kb units, and the corresponding contact map was generated and normalized through the sequential component normalization procedure of SCN (based on the sequential component normalization; https://github.com/koszullab/E_coli_analysis) (Lioy et al., 2018). Contact maps were visualized as logarithmic matrices to aid in visualization.

Ratio of contact maps

The comparison of contact maps was facilitated by displaying their ratio. The ratio was calculated for each point on the map by dividing the number of contacts in one condition by the number of contacts in the other condition. The Log2 of the ratio was then plotted using a Gaussian filter. The color code represented a decrease or increase in contacts in one condition relative to the other (a blue or red signal, respectively); a white signal indicated no change.

Quantification of the range of cis contacts along the chromosome

We used a three-step process adapted from (Lioy et al., 2020) and (Wang et al., 2017) to determine the width of the diagonal in contact maps. To improve the resolution of the Ter limit, we measured the width perpendicularly to the main diagonal.

First, we calculated the median of the contact map and estimated the standard deviation (σ) using a robust statistic, where σ = 1.4826 * mad (mad = median absolute deviation). Next, we used a point connecting algorithm to differentiate significant interactions from background noise. The size of each connected element identified by the "bwlabel" function of MATLAB was determined. We considered a connected element with a size greater than or equal to 30 points to be significant and used the "imclose()" function of MATLAB to fill in the empty points within the connected elements using a diamond shape with a size of 5.

Subsequently, we calculated the width of the primary diagonal for each 5-kb bin of the genome. The range of cis contact was estimated from the width of the primary diagonal by multiplying the number of measured bins by the bin size (5 kb) and dividing by two (since the range is symmetric on both sides of the chromosomal locus being considered). Finally, a boxplot was used to visualize the entire range of cis contacts for all the analyzed chromosome regions.

Directional Index analysis

Directional Index was calculate as described in (Lioy et al., 2018). The directional index is a statistical metric that quantifies the level of upstream or downstream contact bias for a given genomic region (Dixon et al., 2012). This metric is based on a t-test comparison of contact vectors to the left and right of each bin, up to a certain scale. The boundaries between topological domains often generate fluctuating signals that result in a transition in the directional preference. Specifically, for each 5 kb bin, we extracted the contact vector from the correlation matrix between that bin and neighboring bins at regular 5 kb intervals, up to 100 kb, in both left and right directions. At each step, the paired t-test was used to determine whether the strength of interactions was significantly stronger in one direction relative to the other. A threshold of 0.05 was used to assess statistical significance. The directional preferences for each bin along the chromosome were visualized as a bar plot, with positive and negative t-values shown as red and green bars, respectively. To improve the clarity of presentation, bars for bins with t-values below −2 or above 2 (corresponding to a p-value of 0.05) were truncated. Between two identified domains in the contact matrices, the directional preference of bins changed, which was indicated by alternating red and green colors.

Supporting information

eLife Assessment

This paper presents important new data related to the non-canonical SMC complex MukBEF in E. coli, which remains less well understood compared to more canonical SMC complexes in many other bacteria. The authors use a combination of Hi-C and ChIP-seq to demonstrate that MukBEF loads at multiple locations along the chromosome, preferentially in newly replicated DNA, with the MatP/matS system antagonizing MukBEF activity and localizing in terminus proximal regions. Most of the data to support these findings are compelling, with modest concerns raised about the effect sizes of the ChIP-seq data, which could be bolstered with some additional controls.

Reviewer #1 (Public Review):

Summary:

In this manuscript, Seba et al., investigate the mechanism of chromosome organization by the MukBEF complex in E. coli. They use a combination of Hi-C and ChIP analysis to understand the steps of MukBEF regulation: its unloading from DNA (how MukBEF activity is prevented in the terminus regions of the chromosome by MatP), and its loading onto DNA (how DNA replication influences MukBEF association with the chromosome). Seba et al., induce chromosomal rearrangements to flip the sections of the ter region, thus perturbing matS site numbers and position. They find that MukBEF activity is prevented around matS sites, and that higher matS density has greater effect on MukBEF. Separately, using replication mutants and inducible MukBEF expression, they find that MukBEF can associate with the chromosome even in the absence of replication (as seen by the emergence of long-range contacts). However, ChIP data suggests that MukBEF binding to DNA is enriched on newly replicated DNA.

Strengths:

Altogether, this work provides a valuable and comprehensive view of MukBEF-mediated chromosome organization, with insights on the mechanism of the exclusion of MukBEF from the terminus region of the chromosome. The use of the programmed genetic rearrangements is powerful, and allows the authors to provide clear and convincing evidence for MukBEF exclusion from ter by matS sites. It is particularly striking to see that MukBEF can promote long-range contacts even in chromosomal regions between two matS, but the complex is excluded from the matS 'zones'. Experiments using cells blocked for replication show that MukBEF can influence chromosome organization in the absence of replication as well. While previous studies have reported some evidence in support of both of the above conclusions, the experiments described here offer clear and direct demonstration of the same.

Limitations:

A few control experiments are required to strengthen conclusions. Additionally, the discussion section is lacking many references and key papers have not been cited (paragraph 1 of discussion for example has no references). The possibility that SMC-ScpAB and MukBEF can act independent of replication has been suggested previously, but are not cited or discussed. Similarly, there is some evidence for SMC-ScpAB association with newly replicated DNA (PMID 21923769).

Reviewer #2 (Public Review):

Summary:

Chromosome organization in E. coli and related species ('transversal') deviates starkly from the pattern more commonly found in bacteria ('longitudinal'). The underlying mechanisms and the physiological roles, however, are not well understood. This manuscript by Seba et al. investigates the activity and regulation of MukBEF in chromosome folding in E. coli. Using a construct for inducible expression of MukBEF, the authors first demonstrate that the initiation of long-range chromosome contacts (likely by loop extrusion) is not restricted to few positions on the chromosome and rather widely distributed but excluding the replication terminus region. Using ChIP-Seq, the authors show that the distribution of MukBEF over the chromosome is consistent with widely distributed loading and moreover indicate a connection of chromosome folding and DNA replication with newly replicated DNA shower an increased tendency for MukBEF binding. To dissect this further, they then redistribute matS sites on the chromosome by a clever strategy based on large-scale transpositions. The results reveal that matS-free DNA segments undergo MukBEF dependent folding regardless of their position relative to the origin of replication, being consistent with a broad distributed loading of MukBEF. By fine-mapping with smaller transposition events, they show that few matS sites are sufficient to impede MukBEF activity. Surprisingly, however, E. coli and most related genomes harbor many matS sites, which are particularly highly concentrated near the chromosome dimer resolution dif site (Fig. 5).

Strengths:

This is a well-executed and well-presented study. The findings show that the MatP/matS system acts locally and independently of DNA replication to restrict MukBEF in the replication terminus region. Few of the many matS sites are sufficient for MukBEF restriction. The main conclusions of the work are clear and well supported by the data.

Weaknesses:

The biological relevance of MukBEF restriction from the replication terminus region remains unresolved. The authors could speculate on possible functions.

Reviewer #3 (Public Review):

Seba et al. investigate whether chromosomal recruitment of the E. coli SMC complex MukBEF is initiated at a single site, how MukBEF activity is excluded from the replication terminus region, and whether its recruitment and activity depend on DNA replication. Upon induction of MukBEF, the authors find that chromosomal long-range contacts increase globally rather than from a single site. Using large-scale chromosome rearrangements, they show that matS sites can insulate separate areas of high MukBEF activity from each other. This suggests that MukBEF loads at multiple sites in the genome. Finally, the authors propose that MukBEF associates preferentially with newly replicated DNA, based on ChIP-seq experiments after DNA replication arrest.

The conclusions of the paper are mostly well supported by the data. The ratiometric contact analyses and range-of-contact analyses are compelling and nicely show the interplay between MukBEF and its proposed unloader MatP/matS. I particularly enjoyed the chromosome re-arrangement experiments, which lend strong support to the idea that MukBEF activity is independent of a centralized loading site.

The enrichment of MukBEF in newly replicated regions is somewhat less convincing, as the effect sizes are rather small and the background signal is unknown. The conclusion that matS density controls MukBEF activity is appealing, but would likely need additional support from more systematic studies. It is based on a comparison of only two strains (looking at different combinations of three matS sites), and the effect size is small. As it is, differences in matS sequence composition and genomic context cannot be factored out.

Overall, the work is an important advance in our understanding of bacterial chromosome organization. It will be of broad interest to chromosome biologists and bacterial cell biologists.

Author Response

The following is the authors’ response to the original reviews.

Reviewer #1 (Recommendations For The Authors):

1. Experiments regarding the inducible expression of MukBEF: The authors should provide western blots or rt-qPCR for MukBEF expression at 40 min and 2H.

We provide now a western blot of MukB in non-induced and induced conditions as Figure 1-figure supplement 1D.

1. Experiments with RiTer and LiTer constructs:

We agree that the matP mutant may help the reader to compare the effect of the translocation in different backgrounds and have added it to the figure. This strengthens the conclusion that longrange interactions in ter do increase in the absence of matP in a rearranged chromosome, as observed in the WT configuration (Lioy et al., 2018).

b. Additionally, in Fig. 2C, it appears that there is some decrease in long-range interactions in the absence of mukB in ter1 (Riter). Is this a significant change?

The change observed is not significant. The results shown in Fig. 2C have been obtained using a 3C approach, which generated slightly more variability than Hi-C. Furthermore, we measured the range of contacts for the segment corresponding to Ter1 in RiTer (matS12-matS28), in different genetic contexts and different configurations. The results show that this level of variation is not significant (see graph below reporting two independent experiments).

Range of interactions measured on the interval matS12-matS18 in different genetic contexts and different configurations (MG1655 WT(1 and 2), ∆mukB, RiTer, RiTer ∆mukB).

1. Experiments with various matS organizations: These experiments are interesting and an important part of the paper. However, it is rather hard to visualize the chromosome conformations in the strains after transposition. To aid the reader (particularly with panel E), authors can provide schematics of the chromosome conformations and anticipated/ observed chromosomal interactions. Circular interaction plots would be useful here.

We thank the reviewer for this interesting remark; we have tried in the past to represent these interactions using a circular representation (see for example the web site of Ivan Junier; https://treetimc.github.io/circhic/index.html). However, this representation is not trivial to apprehend for nonspecialists, especially in strains with a rearranged chromosome configuration. Nonetheless, we have added graphical circular representations of the chromosome configurations to help the reader.

1. ChIP experiments:

The basal value of the ChIP on the non-replicated sequences (between 0-3.5 Mb for 10 minutes and 0-3 Mb for 20 minutes) is 0.8 and 0.7, respectively, whereas the mean value of the replicated sequence is 1.6 and 1.45. So the enrichment observed for these two points is about 2-fold, not 1.1 and it is 4 fold for t40min. These values were obtained by dividing the number of normalized reads in the ChIP (the number of reads at each position divided by the total number of reads) by the normalized reads of the input. Therefore, the increase in copy number is considered in the calculation. Furthermore, we added a supplementary figure (Figure Sup9) in which we performed a ChIP without tags on synchronized cells, and in this case, we did not observe any enrichment triggered by replication.

b. Authors make a conclusion that MukB loads behind the replication fork. However, the time resolution of the presented experiments is not sufficient to be certain of this. Authors would need to perform more time-resolved experiments for the same.

Reviewer 1 is correct; we attempted to discriminate whether the observed enrichment is (i) associated with the replication fork since we observed a decrease in the center of the enrichment at oriC as the maximum enrichment moves away with the replication fork after 20 and 40 minutes, or (ii) associated with the newly replicated sequence. To investigate this, we attempted to induce a single round of replication by shifting the cells back to 40°C after 10 minutes at 30°C. Unfortunately, replication initiation is not immediately halted by shifting the cells to 40°C, and we were unable to induce a single round of replication. To clarify our conclusions, we modified our manuscript to

“Altogether, these findings indicate that MukBEF is loaded into regions newly replicated either at the replication fork or even further behind it, except in the Ter region from which it would be excluded.”

c. Authors conclude that in the LiTer7 strain, MukB signal is absent from Ter2. However, when compared with the ChIP profiles by eye across panels in A and B, this does not seem to be significant. In the same results sections, authors state that there is a 3-fold increase in MukB signal in other regions. The corresponding graph does not show the same.

Rather than relying solely on the enrichment levels, which can be challenging to compare across different strains due to slight variations in replication levels, we believe there is a clear disruption in this profile that corresponds to the Ter2 sequence. Furthermore, this discontinuity in enrichment relative to the replication profile is also observable in the WT configuration. At T40min, MukB ChIPseq signals halt at the Ter boundary, even though Ter is actively undergoing replication, as evidenced by observations in the input data.

Regarding the fold increase of MukB, Reviewer 1 is correct; we overestimated this enrichment in the text and have now corrected it.

d. Authors should provide western blot of MukB-Flag.

We have added Supplementary Figure 1 D, which contains a Western blot of MukB-Flag.

1. The bioinformatic analysis of matS site distribution is interesting, but this is not followed upon. The figure (Fig 5) is better suited in the supplement and used only as a discussion point.

We acknowledge the reviewer's point, but we used this section to attempt to extend our findings to other bacteria and emphasize the observation that even though a few matS sites are necessary to inhibit MukBEF, the Ter domains are large and centered on dif even in other bacteria.

1. The discussion section is lacking many references and key papers have not been cited (paragraph 1 of discussion for example has no references).

The possibility that SMC-ScpAB and MukBEF can act independent of replication has been suggested previously, but are not cited or discussed. Similarly, there is some evidence for SMC-ScpAB association with newly replicated DNA (PMID 21923769).

We have added references to the suggested paragraph and highlighted the fact that MukBEF's activity independent of replication was already known. However, we believe that the situation is less clear for SMC-ScpAB in B. subtilis or C. crescentus. In a similar manner, we found no clear evidence that SMCScpAB is associated with newly replicated DNA in the referenced studies.

To clarify and enrich the discussion section, we have added a paragraph that provides perspective on the loading mechanisms of SMC-ScpAB and MukBEF.

1. There are minor typographical errors that should be corrected. Some are highlighted here:

a. Abstract: L5: "preferentially 'on' instead of 'in'"

b. Introduction: Para 1 L8: "features that determine"

c. Introduction: Para 2 L1: please check the phrasing of this line

d. Results section 2: L1: Ter "MD" needs to be explained

e. Page 8: Para 2: L6: "shows that 'a'"

g. Page 13: Para 2: "MukBEF activity...". This sentence needs to be fixed.

i. Figure 4: "input" instead of "imput"

We thank Reviewer 1 for pointing out all these grammatical or spelling mistakes. We have corrected them all.

f. Page 12: Para 2: "Xer" instead of "XDS"? *We added a reference to clarify the term.

h. Methods: ChIP analysis: Authors state "MatP peaks", however, reported data is for MukB

This description pertains to the matP peak detection shown in Supplementary Figure 3. We have incorporated this clarification into the text.

j. Supplementary figure legends need to be provided (currently main figure legends appear to be pasted twice)

Supplementary figure legends are provided at the end of the manuscript, and we have edited the manuscript to remove one copy of the figure legends.

k. Authors should ensure sequencing data are deposited in an appropriate online repository and an accession number is provided.

We waited for the appropriate timing in the editing process to upload our data, which we have now done. Additionally, we have added a data availability section to the manuscript, including sequence references on the NCBI.

Reviewer #2 (Recommendations For The Authors):

The authors largely avoid speculation on what might be the physiological relevance of the exclusion of MukBEF (and Smc-ScpAB) from the replication termination region (and the coordination with DNA replication). At this stage it would be helpful to present possible scenarios even if not yet supported by data. The authors should for example consider the following scenario: loop extrusion of a dif site in a chromosome dimer followed by dimer resolution by dif recombination leads to two chromosomes that are linked together by MukBEF (equivalent to cohesin holding sister chromatids together in eukaryotes but without a separase). This configuration (while rare) will hamper chromosome segregation. Is MatP particularly important under conditions of elevated levels of chromosome dimers? Could this even be experimentally tested? Other scenarios might also be entertained.

Even though we prefer to avoid speculations, we agree that we may attempt to propose some hypotheses to the reader. To do so, we have added a few sentences at the end of our discussion. “We may speculate, based on in vitro observations (Kumar et al., 2022), that MukBEF could interfere with TopIV activity and delay potential chromosome decatenation. Another possibility is that chromosome dimers resolved at the dif site may become trapped in loops formed by MukBEF, thus delaying segregation. But none of these possible scenarios are supported by data yet, and a major challenge for the future is to determine whether and how MukBEF may interfere with one or both of these processes.”

The manuscript text is well written. However, the labeling of strains in figures and text is sometimes inconsistent which can be confusing (LiTer Liter liter; e.g Riter Fig 2C). For consistency, always denote the number of matS sites in LiTer strains and also in the RiTer strain. The scheme denoting LiTer and RiTer strains should indicate the orientation of DNA segments so it is clear that the engineering does not involve inversion (correct?). Similarly: Use uniform labelling for time points: see T40mn vs 40mn vs T2H vs 2H

We have reviewed the manuscript to standardize our labeling. Additionally, we have included a schema in Figure 2, indicating the matS numbers at the Ter border to emphasize that the transposition events do not involve inversion.

matS sites do not have identical sequences and bind different levels of MatP (suppl fig 3). Does this possibly affect the interpretation of some of the findings (when altering few or only a single matS site). Maybe a comment on this possibility can be added.

We agree with the referee; we do not want to conclude too strongly about the impact of matS density, so we have added this sentence at the end of the section titled 'matS Determinants to Prevent MukBEF Activity':

“Altogether, assuming that differences in the matS sequences do not modify MatP's ability to bind to the chromosome and affect its capacity to inhibit MukBEF, these results suggested that the density of matS sites in a small chromosomal region has a greater impact than dispersion of the same number of matS sites over a larger segment”

Figure 5: show selected examples of matS site distribution in addition to the averaged distribution (as in supplemental figure)?

Figure 5 shows the median of the matS distribution based on the matS positions of 16 species as displayed in the supplementary figure. We believe that this figure is interesting as it represents the overall matS distribution across the Enterobacterales, Pasteurellales, and Vibrionales.

How do authors define 'background levels' (page 9)in their ChIP-Seq experiments? Please add a definition or reword.

We agree that the term 'background level' here could be confusing, so we have modified it to 'basal level' to refer to the non-replicating sequence. The background level can be observed in Supplementary Figure 9 in the ChIP without tags, and, on average, the background level is 1 throughout the entire chromosome in these control experiments.

This reviewer would naively expect the normalized ChIP-Seq signals to revolve around a ratio of 1 (Fig. 4)? They do in one panel (Figure 4B) but not in the others (Figure 4A). Please provide an explanation.

We thank the referee for this pertinent observation. An error was made during the smoothing of the data in Figure 4A, which resulted in an underestimation of the input values. This mistake does not alter the profile of the ChIP (it's a division by a constant) and our conclusions. We provide a revised version of the figure.

Inconsistent axis labelling: e.g Figure 4

Enterobacterals should be Enterobacterales (?)

KB should be kb

MB should be Mb

Imput should be Input

FlaG should be Flag

We have made the suggested modifications to the text.

'These results unveiled that fluorescent MukBEF foci previously observed associated with the Ori region were probably not bound to DNA' Isn't the alternative scenario that MukBEF bound to distant DNA segments colocalize an equally likely scenario? Please rephrase.

Since we lack evidence regarding what triggers the formation of a unique MukB focus associated with the origin and what this focus could represent, we have removed this sentence.

Reviewer #3 (Recommendations For The Authors):

The text is well-written and easy to follow, but I would suggest several improvements to make things clearer:

1. Many plots are missing labels or legends. (I) All contact plots such as Fig. 1C should have a color legend. It is not clear how large the signal is and whether the plots are on the same scale. (II)

As indicated in the materials and methods section, the ratio presented on this manuscript was calculated for each point on the map by dividing the number of contacts in one condition by the number of contacts in the other condition. The Log2 of the ratio was then plotted using a Gaussian filter.

1. Genotypes and strain names are often inconsistent. Sometimes ΔmukB, ΔmatP, ΔmatS is used, other times it is just mukB, matP, matS; There are various permutations of LiTer, Liter, liter etc.

These inconsistencies have been corrected.

1. The time notation is unconventional. I recommend using 0 min, 40 min, 120 min etc. instead of T0, T40mn, T2H.

As requested, we have standardized and used conventional annotations.

1. A supplemental strain table listing detailed genotypes would be helpful.

A strain table has been added, along with a second table recapitulating the positions of matS in the different strains.

1. Fig. 1A: Move the IPTG labels to the top? It took me a while to spot them.

We have moved the labels to the top of the figure and increased the font size to make them more visible.

1. Fig 1C: Have these plots been contrast adjusted? If so, this should be indicated. The background looks very white and the transitions from diagonal to background look quite sharp.

No, these matrices haven't been contrast-adjusted. They were created in MATLAB, then exported as TIFF files and directly incorporated into the figure. Nevertheless, we noticed that the color code of the matrix in Figure 3 was different and subsequently adjusted it to achieve uniformity across all matrices.

7, Fig 1C: What is the region around 3 Mb and 4 Mb? It looks like the contacts there are somewhat MukBEF-independent.

The referee is right. In the presence of the plasmid pPSV38 (carrying the MukBEF operon or not), we repeatedly observed an increase of long range contacts around 3 Mb. The origin of these contacts is unknown.

1. Fig 1D: Have the log ratios been clipped at -1 and 1 or was some smoothing filter applied? I would expect the division of small and noisy numbers in the background region to produce many extreme values. This does not appear to be the case.

The referee is right, dividing two matrices generates a ratio with extreme values. To avoid this, the Log2 of the ratio is plotted with a Gaussian filter, as described before (Lioy et al., 2018).

1. Fig 1E: I recommend including a wild-type reference trace as a point of reference.

We have added the WT profile to the figure.

1. Fig 2: I feel the side-by-side cartoon from Supplemental Fig. 2A could be included in the main figure to make things easier to grasp.

We added a schematic representation of the chromosome configuration on top of the matrices to aid understanding.

1. Fig. 2C: One could put both plots on the same y-axis scale to make them comparable.

We have modified the axes as required.

1. Fig. 3C: The LiTer4 ratio plot has two blue bands in the 3-4.5 Mb region. I was wondering what they might be. These long-range contacts seem to be transposition-dependent and suppressed by MatP, is that correct?

The referee is right. This indicates that in the absence of MatP, one part of the Ter was able to interact with a distal region of the chromosome, albeit with a low frequency. The origin is not yet known.

1. Fig. 3E: It is hard to understand what is a strain label and what is the analyzed region of interest. The plot heading and figure legend say Ter2 (but then, there are different Ter2 variants), some labels say Ter, others say Ter2, sometimes it doesn't say anything, some labels say ΔmatS or ΔmatP, others say matS or matP, and so on.

We have unified our notation and add more description on the legend to clarify this figure :

“Ter” corresponds to the range of contacts over the entire Ter region, in the WT strain (WT Ter) or in the ΔmatP strain (ΔmatP Ter). The column WT matSX-Y corresponds to the range of contacts between the designated matS sites in the WT configuration. This portion of the Ter can be compared with the same Ter segment in the transposed strain (Ter2). Additionally, the matS20-28 segment corresponds to Ter2 in LiTer9, just as matS22-28 corresponds to Ter2 in LiTer7, and matS25-28 to Ter2 in LiTer4. The range of contacts of this segment was also measured in a ΔmatP or ΔmatS background.”

1. Fig. 4 and p.9: "Normalized ChIP-seq experiments were performed by normalizing the quantity of immuno-precipitated fragments to the input of MukB-Flag and then divide by the normalized ChIP signals at t0 to measure the enrichment trigger by replication."

This statement and the ChIP plots in Fig. 4A are somewhat puzzling. If the data were divided by the ChIP signal at t0, as stated in the text, then I would expect the first plot (t0) to be a flat line at value 1. This is not the case. I assume that normalized ChIP is shown without the division by t0, as stated in the figure legend.

The referee is right. This sentence has been corrected, and as described in the Methods section, Figure 4 shows the ChIP normalized by the input.

If that's true and the numbers were obtained by dividing read-count adjusted immunoprecipitate by read-count adjusted input, then I would expect an average value of 1. This is also not the case. Why are the numbers so low? I think this needs some more details on how the data was prepared.

The referee is right; we thank him for this remark. Our data are processed using the following method: the value of each read is divided by the total number of reads. A sliding window of 50 kb is applied to these normalized values to smooth the data. Then, the resulting signal from the ChIP is divided by the resulting signal from the input. This is what is shown in Figure 4. Unfortunately, for some of our results, the sliding window was not correctly applied to the input data. This did not alter the ChIP profile but did affect the absolute values. We have resolved this issue and corrected the figure.

Another potential issue is that it's not clear what the background signal is and whether it is evenly distributed. The effect size is rather small. Negative controls (untagged MukB for each timepoint) would help to estimate the background distribution, and calibrator DNA could be used to estimate the signal-to-background ratio. There is the danger that the apparent enrichment of replicated DNA is due to increased "stickiness" rather than increased MukBEF binding. If any controls are available, I would strongly suggest to show them.

To address this remark, a ChIP experiment with a non-tagged strain under comparable synchronization conditions has been performed. The results are presented as Supplementary Figure 9; they reveal that the enrichment shown in Figure 4 is not attributed to nonspecific antibody binding or 'stickiness’.

1. Fig. 4A, B: The y-axes on the right are unlabeled and the figure legends mention immunoblot analysis, which is not shown.

We labeled the y-axes as 'anti-Flag ChIP/input' and made corrections to the figure legend.

1. Fig. 4B: This figure shows a dip in enrichment at the Ter2 region of LiTer7, which supports the authors' case. Having a side-by-side comparison with WT at 60 min would be good, as this time point is not shown in Fig. 4A.

Cell synchronization can be somewhat challenging, and we have observed that the timing of replication restart can vary depending on the genetic background of the cells. This delay is evident in the case of LiTer7. To address this, we compared LiTer7 after 60 minutes to the wild type strain (WT) after 40 minutes of replication. Even though the duration of replication is 20 minutes longer in LiTer7, the replication profiles of these two strains under these two different conditions (40 minutes and 60 minutes) are comparable and provide a better representation of similar replication progression.

1. Fig. 4C: Highlighting the position of the replication origin would help to interpret the data.

We highlight oriC position with a red dash line

1. Fig. 4C: One could include a range-of-contact plot that compares the three conditions (similar to Fig. 1E).

We have added this quantification to Supplemental Figure 8

1. Supplemental Fig. 2A: In the LiTer15 cartoon, the flanking attachment sites do not line up. Is this correct? I would also recommend indicating the direction of the Ter1 and Ter2 regions before and after recombination.

In this configuration, attB and attR, as well as attL and attB', should be aligned but the remaining attR attL may not. We have corrected this misalignment. To clarify the question of sequence orientation, we have included in the figure legend that all transposed sequences maintain their original orientation.

1. Supplemental Fig. 3: One could show where the deleted matS sites are.

We added red asterisks to the ChIP representation to highlight the positions of the missing matS.

1. Supplemental Fig. 3B: The plot legend is inconsistent with panel A (What is "WT2")?

We have corrected it.

1. Supplemental Fig. 3C: The E-value notation is unusual. Is this 8.9 x 10^-61?

The value is 8.9 x 10-61; we modified the annotation.

1. Abstract: "While different features for the activity of the bacterial canonical SMC complex, SmcScpAB, have been described in different bacteria, not much is known about the way chromosomes in enterobacteria interact with their SMC complex, MukBEF."

Could this be more specific? What features are addressed in this manuscript that have been described for Smc-ScpAB but not MukBEF? Alternatively, one could summarize what MukBEF does to capture the interest of readers unfamiliar with the topic.

We modified these first sentences.

1. p.5 "was cloned onto a medium-copy number plasmid under control of a lacI promoter" Is "lacI promoter" correct? My understanding is that the promoter of the lacI gene is constitutive, whereas the promoter of the downstream lac operon is regulated by LacI. I would recommend providing an annotated plasmid sequence in supplemental material to make things clearer.

We modified it and replaced “ lacI promoter” with the correct annotation, pLac.

1. p. 5 heading "MukBEF activity does not initiate at a single locus" and p. 6 "Altogether, the results indicate that the increase in contact does not originate from a specific position on the chromosome but rather appears from numerous sites". Although this conclusion is supported by the follow-up experiments, I felt it is perhaps a bit too strong at this point in the text. Perhaps MukBEF loads slowly at a single site, but then moves away quickly? Would that not also lead to a flat increase in the contact plots? One could consider softening these statements (at least in the section header), and then be more confident later on.

We used 'indicate' and 'suggesting' at the end of this results section, and we feel that we have not overreached in our conclusions at this point. While it's true that we can consider other hypotheses, we believe that, at this stage, our suggestion that MukBEF is loaded over the entire chromosome is the simplest and more likely explanation.

1. p.7: "[these results] also reveal that MukBEF does not translocate from the Ori region to the terminus of the chromosome as observed with Smc-ScpAB in different bacteria."

This isn't strictly true for single molecules, is it? Some molecules might translocate from Ori to Ter. Perhaps clarify that this is about the bulk flux of MukBEF?

At this point, our conclusion that MukBEF does not travel from the ori to Ter is global and refers to the results described in this section. However, the referee is correct in pointing out that we cannot exclude the possibility that in a WT configuration (without a Ter in the middle of the right replicore), a specific MukBEF complex can be loaded near Ori and travel all along the chromosome until the Ter. To clarify our statement, we have revised it to 'reveal that MukBEF does not globally translocate from the Ori region to the terminus of the chromosome.' This change is intended to highlight the fact that we are drawing a general conclusion about the behavior of MukBEF and to facilitate its comparison with Smc-ScpAB in B. subtilis.

1. p. 10: The section title "Long-range contacts correlate with MukBEF binding" and the concluding sentence "Altogether, these results indicate that MukBEF promotes long-range DNA contacts independently of the replication process even though it binds preferentially in newly replicated regions" seem to contradict each other. I would rephrase the title as "MukBEF promotes long-range contacts in the absence of replication" or similar.

We agree with this suggestion and have used the proposed title.

1. p. 13: I recommend reserving the name "condensin" for the eukaryotic condensin complex and using "MukBEF" throughout.

We used MukBEF throughout.

Version 2

Abstract

Abstract

Structural maintenance of chromosomes (SMC) complexes share conserved structures and serve a common role in maintaining chromosome architecture. In the bacterium Escherichia coli, the SMC complex MukBEF is necessary for rapid growth and the accurate segregation and positioning of the chromosome, although the specific molecular mechanisms involved are still unknown. Here we used a number of in vivo assays to reveal how MukBEF controls chromosome conformation and how the MatP/matS system prevents MukBEF activity. Our results indicate that the loading of MukBEF occurs preferentially on newly replicated DNA, at multiple loci on the chromosome where it can promote long-range contacts in cis even though MukBEF can promote long-range contacts in the absence of replication. Using HiC and ChIP-seq analyses in strains with rearranged chromosomes, the prevention of MukBEF activity increases with the number of matS sites and this effect likely results from the unloading of MukBEF by MatP. Altogether, our results reveal how MukBEF operates to control chromosome folding and segregation in E. coli.

Introduction

SMC complexes play key roles in many processes involved in chromosome management, from genome maintenance, interphase chromatin organization, sister chromatids alignment, chromosome folding and condensing, to DNA recombination at specific stages of the cell cycle (Yatskevich et al., 2019). A general model for SMC complex activity relies on their properties to bridge DNA elements and by doing so build DNA loops in cis and hold together the sister chromatids in trans; to do that, they bind DNA and processively extrude a DNA loop in a ATPase-driven loop extrusion, thereby compacting and organizing DNA (Davidson et al., 2019; Ganji et al., 2018; Kong et al., 2020). Yet many molecular features that determine the activity of the various SMC complexes are still unclear and it is not known whether they work using the same basal mechanisms (Bürmann et al., 2021; Davidson et al., 2019; Pradhan et al., 2022). For example, it is not yet clear whether i) the activity of SMC complexes is mediated by a single complex or if it involves cooperation between several complexes that organize into dimers or even oligomers (Badrinarayanan et al., 2012b; Hassler et al., 2018), ii) the DNA extrusion involves a topological or nontopological mechanism, i.e. does the DNA pass through the SMC ring topologically (Bürmann et al., 2021; Davidson et al., 2019; Pradhan et al., 2022), iii) the activity of SMC complexes relies on the folding capacities of the SMC coiled-coil arms facilitating large-scale conformational changes (Bürmann et al., 2019), and iv) multiple complexes that encounter one another on the same DNA in living cells bypass each other or collide (Anchimiuk et al., 2021; Brandão et al., 2021). Cohesin and Condensin are the most characterized SMC complexes found in many eukaryotes. In bacteria, three different forms of SMC-like complexes defined as bacterial condensins have been identified, Smc-ScpAB, MukBEF, MksBEF; they are considered functionally related to condensins as they are thought to compact chromosomes and facilitate the segregation of sister chromosomes (Lioy et al., 2018, 2020; Marbouty et al., 2015). In bacteria, Smc-ScpAB represents the most highly conserved complex, while MukBEF and MksBEF represent diverged SMC complexes (Cobbe and Heck, 2004; Yoshinaga and Inagaki, 2021).

Eukaryotic and prokaryotic SMC complexes are composed of at least 5 subunits: two Smc subunits, a kleisin subunit, and two additional subunits (referred to as kite and hawk subunits, depending on the type of SMC complex). Smc proteins associate with the kleisin protein to form a ring-shaped ATPase assembly. Additional subunits associate with this tripartite complex: the “Kite” family associates with bacterial and archaeal SMC complexes and also with the eukaryotic SMC5/6 complex while the “Hawk” family interacts with condensin and cohesin (Yatskevich et al., 2019). These additional subunits are thought to be required for the activity and to differentiate functions; for example, while condensin I and II share the same pair of Smc proteins, the difference in the subunit composition specifies their spatiotemporal dynamics and functional contributions to mitotic chromosome assembly (Hirota et al., 2004; Kong et al., 2020).

SMC function and dynamics on DNA requires additional auxiliary proteins (Baxter et al., 2019). More specifically, its loading on DNA and unloading of the DNA may depend on specific factors, at specific sites. For example, in Bacillus subtilis, the segregation ParB protein bound to parS site directly bind the Smc subunit; the ParB clamp presumably present DNA to the SMC complex to initiate DNA loop extrusion (Gruber and Errington, 2009; Sullivan et al., 2009; Wang et al., 2015, 2017). Upon translocation, the site-specific recombinase XerD bound to its binding site unloads SMC complexes in the terminus region of the chromosome and this process is thought to involve specific interactions between the different components (Karaboja et al., 2021).

MukBEF was the first SMC complex identified. In E. coli, MukBEF is thought to be required for chromosome segregation as muk mutants present many anucleate cells or mis-segregated chromosomes (Niki et al., 1991). How MukBEF may promote chromosome segregation and organization has remained elusive for a long time and is still not clear (Nolivos and Sherratt, 2014). The effect of the MukBEF complex in E. coli appears to be radically different from that of SMC in B. subtilis or other bacteria. Instead of aligning the chromosome arms from a centromere-like locus, MukBEF promotes DNA contacts in the megabase range within each replication arm (Lioy et al., 2018, 2020). MukBEF promotes long range interactions along the chromosome except in the 800-kb long terminal domain (Ter) where MatP prevents its activity (Lioy et al., 2018). Under conditions of increased chromosome occupancy of MukBEF, the E. coli chromosome appears to be organized as a series of loops around a thin (<130 nm) MukBEF axial core (Mäkelä and Sherratt, 2020). Whether MukBEF is loaded at a particular locus is still an open question. The complete atomic structure of MukBEF in complex with MatP and DNA has been determined by electron cryomicroscopy (Bürmann et al., 2021); it contains also the MukBEF binding partner AcpP protein (Prince et al., 2021). It revealed that the complex binds two distinct DNA double helices reminiscent of the arms of an extruded loop, MatP- bound DNA threads through the MukBEF ring, while the second DNA is clamped by MukF, MukE, and the MukB ATPase heads. The presence of MatP in the complex together with its ability to prevent MukBEF activity prompted authors to propose that MatP might be an unloader of MukBEF (Bürmann et al., 2021).

Here we have performed a number of experiments using different in vivo approaches to further characterize how MukBEF contributes to chromosome management in E. coli. We have used ChIP-seq and Chromosome Conformation Capture (3C-seq and HiC) experiments to study how MukBEF is loaded on the chromosome and promotes long range DNA contacts. By using strains with various chromosome configurations obtained by programmed genetic rearrangements, we have explored how MukBEF activity proceeds along the chromosome and how MatP bound to matS sites prevents its activity. Our results together with comparative genomics analyses allow us to address the biological significance of multiple matS sites in the Ter region of chromosomes in Enterobacteria and of the absence of MukBEF activity.

Results

MukBEF activity does not initiate at a single locus

In order to characterize how MukBEF interacts with the chromosome and initiate its activity of long-range contacts, it was necessary to set-up a system to reveal using HiC the appearance and spread of long-range contacts along the chromosome upon MukBEF synthesis. The rationale was based on previous findings showing that long-range DNA contacts within replication arms, outside Ter, result from MukBEF activity. If MukBEF loads at a specific locus as observed for Smc-ScpAB at parS sites, we would expect to detect the appearance and spreading of long-range contacts from this site upon MukBEF synthesis. By contrast, if MukBEF loads stochastically or at multiple loci on the chromosome, long-range contacts should occur at multiple sites.

The analysis of MukBEF activity required an efficient system to control its activity. To conditionally synthesize MukBEF, the mukBEF operon was cloned onto a medium-copy number plasmid under control of a pLac promoter and introduced in a mukF mutant. In the absence of inducer, as observed for a mukF mutant, no growth was detected at 37°C and the amount of anucleated cells at 22°C was similar to that of the mukF mutant (13% versus 15%). Induced expression of mukBEF functionally complemented the absence of MukF, restoring growth at 37°C in LB (Figure 1A) and accurately segregating the chromosome at 22°C, as evidenced by the low amount of anucleate cells (less than 2%) observed in the presence of the inducer (Figure 1B).

MukBEF activity detected along the chromosome.

(A) Complementation of the ΔmukF mutant. Serial dilutions (10-1 to 10-5) of an exponential culture of MG1655 (WT), MG1655 ΔmukF ppSV38, and MG1655 ΔmukF ppSV38::mukBEF were plated on media with (Right) or without (Left) inducer (IPTG). Plates were incubated at a permissive temperature of 22°C (top) or at a non-permissive temperature of 37°C (bottom).

(B) : Percentage of anucleate cells (blue bars) in wild type (WT) and mukF mutant strains, complemented or not by the plasmid ppsV38::mukBEF, grown in minimal medium at 22°C with or without IPTG. Anucleate cells were identified and counted after DAPI staining. The histograms and error bars represent the means and standard deviations from at least three independent experiments.

(C) : Normalized Hi-C contact maps (5Kb bin resolution) obtained from a ΔmukF strain complemented with ppsV38::mukBEF after different induction times (0 min, 40 min, 120 min). Cells were grown in permissive conditions at 22°C in minimal medium. The X and Y axes represent genomic coordinates in megabases (Mb). Dashed lines indicate the Ter position.

(D) Ratio of normalized contact maps of ΔmukF ppSV38::mukBEF grown in the presence and absence of ITPG, represented in the left panel for 40 minutes of growth. The right panel shows the ratio of normalized contact maps for 2 hours of growth with and without ITPG. A decrease or increase in contacts in the induced condition compared with the non-induced condition is represented with a blue or red color, respectively. The black line represents a schematic chromosome, with the Ter domain highlighted in light blue. Dashed lines indicate the Ter position.

(E) Quantification of the range of cis contacts of chromosomal loci along the chromosome of a wt strain grown at 22°C (green line) and of a ΔmukF derivative carrying ppsV38::mukBEF grown at 22°C under four different conditions: light blue (without ITPG), red (after 20 minutes of ITPG), yellow (after 40 minutes of ITPG), and purple (after 2 hours of ITPG). Dashed lines indicate the Ter position.

To determine how MukBEF activity initiates in the E. coli chromosome, we induced the expression of mukBEF and monitored the appearance of long-range DNA contacts at different times after induction. MukBEF synthesis was monitored by western blot (Figure 1-figure supplement 1A, D). HiC contact maps were established 20 min, 40 min and 2 hours after induction and compared to that obtained in the absence of induction (Figure 1C – figure supplement 1B, C): long-range DNA contacts were hardly visible after 20 min (Figure 1C -figure supplement 1A), were readily observed after 40 min and reached after 2 hours a level similar to that observed in wild-type strains (figure supplement 1C). The ratio of normalized contact maps of the induced strain at different time points to the non-induced strain allowed to visualize the presence of long-range contacts all over the chromosome except in Ter (Figure 1D).

The range of DNA contacts along the chromosome was quantified by measuring the width of the diagonal perpendicular to it, using an adapted version of the quantification method developed before (Wang et al., 2017 ; Lioy et al., 2020). The plot obtained (Figure 1E) allows to estimate the effect of MukBEF activity on all loci of the E. coli genome. This confirmed the increase in the range of contacts following induction of MukBEF predicted from the ratio of normalized contact maps, with 40 minutes necessary to observe a significant effect and 2 hours to reach the maximum range of contact, as observed in the wild-type strain (Lioy et al., 2020) (Figure1-Figure supplement 1). As expected, Ter region was not affected by the induction of MukBEF because of the presence of MatP bound to matS sites.

Altogether, the results indicate that the increase of contacts does not originate from a specific position on the chromosome but rather appears from numerous sites, suggesting that the activity of MukBEF does not initiate at a single locus but rather from multiple loci along the chromosome.

MukBEF activity initiates in different regions of the E. coli genome

The MatP/matS system has been shown to prevent MukBEF activity in Ter (Lioy et al. 2018). We took advantage of this property to unravel mechanistic aspects of MukBEF activity and address the following issues: could MukBEF interacts with a region flanked by different Ter segments and could MukBEF operate from different regions of the chromosome. To answer these questions, we used bacteriophage λ site-specific recombination as described before (Thiel et al., 2012) to perform large transpositions of a segment of the right (“RiTer”) and left (“LiTer”) replichores at different loci (figure supplement 2A), in Ter, and analyze MukBEF activity in the resulting strain (Figure 2A).

MukBEF activity initiates at different regions of the E.coli chromosome.

(A) (left) Normalized 3C-seq 5kb bin contact map of strain RiTer11 and (right) normalized Hi-C 5Kb bin contact map of strain LiTer15. A circular representation of the chromosome for each strain has been drawn on top of each matrix, with the Ter sequence highlighted in blue and the matS delineating each region. Below the matrices, a linear map is provided, and dashed lines indicate the positions of the two Ter sequences on each matrix.

(B) Ratio of normalized contact maps of RiTer11 strain on RiTer11 ΔmukB strain (left), or RiTer11 on RiTer11 ΔmatP (middle), and LiTer15 on LiTer15 ΔmukB strain (right). A decrease or increase in contacts in the transposed cells compared to the transposed mutant cells is shown in blue or red color, respectively. Dashed lines indicate the Ter position.

(C) The graph shows the quantification of the Hi-C diagonal width for loci along the chromosome of the RiTer11 strain (left panel, blue line) and the RiTer11 ΔmukB strain (left panel, red line), as well as LiTer15 (right panel, blue line) and LiTer15 ΔmukB strain (right panel, red line). Schematic representations of the two Ter segments are indicated below the graph, highlighting the fact that the range of contacts decreased on all Ter fragments. Dashed lines indicate the Ter position.

The transposition in the RiTer11 configuration results in a region of 450 kb devoid of matS, located between two Ter segments of 378 kb and 437 kb, which contain 11 matS and 17 matS respectively (Figure 2-figure supplement 2A). 3C-seq experiments were performed with this strain (“RiTer11”) and on the same strain deleted for matP or for mukF (Figure 2A- figure supplement 2B). The ratio of normalized contact maps of the RiTer11 strain to the RiTer11 ΔmatP allowed to visualize the reduction of long-range contacts in the two Ter segments in the presence of MatP (Figure 2B). On the other hand, the ratio of normalized contact maps of the RiTer11 strain to the RiTer11 ΔmukB allowed to visualize the increase of long-range contacts between and outside the two Ter segments in the presence of MukBEF. Remarkably, long-range DNA contacts specific of MukBEF activity were observed over the chromosome except in the two Ter segments. The range of DNA contacts along the chromosome were quantified by measuring the width of the diagonal in the two strains; these plots (Figure 2C) allow to estimate the effect of MukBEF activity on all loci inside and outside Ter segments. These results indicated that MukBEF can operate to form long-range contacts between the two Ter segments and that MatP can prevent MukBEF activity when segments carrying matS sites are moved in different locations of the genome.

Similar experiments were performed with a strain carrying a transposed segment from the left replichore in the Ter MD (LiTer 15). In this configuration, a region of 1085 kb devoid of matS, is flanked by two Ter segments of 427 kb and 374 kb containing 13 matS and 15 matS respectively (Figure 2- Figure supplement 2A). HiC experiments were performed on this strain (“LiTer”) (Figure 2A). As observed with the RiTer11 strain, long-range DNA contacts specific of MukBEF activity were observed over the chromosome except in the two Ter segments. The ratio of normalized contact maps of the LiTer15 strain to the LiTer15 ΔmukB allowed to visualize the increase of long-range contacts outside the two Ter segments in the presence of MukBEF (Figure 2B and 2C- figure supplement 2C).

Altogether, these results demonstrate that MukBEF activity can be initiated from multiple regions of the chromosome, in the Right, Left or Ori regions, and that Ter segments cannot insulate DNA segments devoid of matS sites. They also reveal that MukBEF does not globally translocate from the Ori region to the terminus of the chromosome as observed with Smc-ScpAB in different bacteria. Furthermore, the results indicated that Ter segments carrying 17 and 11 matS can prevent MukBEF activity and restrict DNA contacts.

matS determinants to prevent MukBEF activity

The method described above, splitting Ter in two parts, revealed that two parts of Ter can prevent MukBEF activity. Thus, by varying the way Ter is split in two parts, one should be able to identify matS determinants required to affect MukBEF activity.

Two previous studies differ slightly in the prediction for the number and consensus sequence of matS (Mercier et al., 2008; Nolivos et al., 2016).To clarify this, three independent ChIP-seq experiments were performed, revealing 28 matS and a newly derived consensus sequence (Figure 3- figure supplement 3). To examine the requirements for the inhibition of MukBEF by MatP, a 1 Mb region of the left replichore was transposed at different positions into the Ter domain, thus dividing Ter into two parts (referred to as Ter1 and Ter2, Ter1 comprising the dif site and the terminus of replication) of different sizes and carrying different number of matS (Figure 3A). Three transpositions were performed generating Ter2 domains of 253, 209 and 136 kb containing 9, 7 or 4 matS, respectively (strains LiTer9, LiTer7, LiTer4). We used Hi-C to test the ability of these Ter2 regions to inhibit MukBEF (LiTer9 in Figure 3-figure supplement 4A, LiTer7 and LiTer4 in Figure 3B). Long-range DNA interactions were readily observed on both regions flanking Ter2 in both three strains, and long-range contacts are affected in the Ter2 segment, even when only 4 matS were present (Figure 3B). Similar HiC experiments were performed with the same strains deleted for matP (figure supplement 4B). The ratio of normalized contact maps of the LiTer4 strains to the corresponding matP derivatives (Figure 3C) clearly revealed that long-range contacts were limited in Ter2 of the different strains. Remarkably, the 136 kb region carrying 4 matS sites in LiTer4 was sufficient to decrease long-range contacts promoted by MukBEF activity.

The number and distribution of matS sites have different effects on MukBEF Inhibition.

(A) The figure shows a schematic representation of different transpositions with the number of matS sites located on each Ter region. The three att sites are integrated into the chromosome in the same orientation, and attL/attR are fixed positions on all LiTer transpositions. attB is inserted at different positions into the Ter allowing, upon transposition, the division of the Ter domain into two subdomains, Ter1 and Ter2, containing different numbers of matS sites. The number of matS sites is indicated for the different transpositions, and the distribution of matS on the Ter2 segment is indicated in the schematic for the three transposed strains, LiTer4, LiTer7, and LiTer9.

(B) Normalized Hi-C contact map with 5Kb bin resolution of the transposed strains LiTer7, LiTer4, and LiTer4 ΔmatS28. The position of the different Ter regions is highlighted below the matrix, and by dashed lines on the matrix.

(C) Ratio of normalized 5Kb bin contact maps for the different transpositions compared to the matP mutant on the same genetic organization. The position of the different Ter regions is highlighted below the ratio, and by dashed lines on the ratio.

(D) Quantification of the Hi-C diagonal width for loci along the chromosome for the transposed strains LiTer7, LiTer4, the derivative mutant LiTer4ΔmatS28, and LiTer4ΔmatP. The schematic map below represents the LiTer7 configuration. The position of the different Ter regions is highlighted below the graph, and by dashed lines on the graph.

(E) This panel quantifies the range of contacts in the Ter2 region or in the corresponding sequence on the wild-type (WT) configuration. Boxplot representations are used, indicating the median (horizontal bar), the 25th and 75th percentiles (open box), and the rest of the population. “Ter” corresponds to the range of contacts over the entire Ter region, in the WT strain (WT Ter) or in the ΔmatP strain (ΔmatP Ter). The column WT matSX-Y corresponds to the range of contacts between the designated matS sites in the WT configuration. This portion of the Ter can be compared with the same Ter segment in the transposed strain (Ter2). Additionally, the matS20-28 segment corresponds to Ter2 in LiTer9, just as matS22-28 corresponds to Ter2 in LiTer7, and matS25-28 to Ter2 in LiTer4. The range of contacts of this segment was also measured in a ΔmatP or ΔmatS background.

The ability of different Ter segments to inhibit the formation of long-range DNA contacts was quantified by measuring the range of contacts in the different Ter segments (Figure 3D and E). The median range of contact for Ter in the WT condition is 263 +/- 24kb. Remarkably, no significant differences in the range of contacts (275 +/- 9kb) were observed in strain Liter9 for the Ter2 segment carrying 9 matS; it is noteworthy that this region carrying 9 matS sites shows the same range of contacts (280 +/- 35kb) in the wt configuration. This result shows that a transposed 253-kb region carrying 9 matS prevent MukBEF activity as much as the same segment present in the 838-kb long Ter MD.

To further characterize how MatP/matS can prevent MukBEF activity, we measured the range of contacts in Ter2 segments carrying 7 or 4 matS sites. For the 209-kb segment carrying 7 matS sites, the range of contact was increased to 330 +/- 9kb while it increased to 352 +/- 12kb for the 136-kb segment with 4 matS. Values obtained for these segments when present as part of the WT Ter were slightly lower (Figure 3E). These results indicate that segments carrying 7 or 4 matS affect MukBEF activity, even though at a lower level than a 250-kb segment with 9 matS sites.

The four matS sites present in the 136-kb Ter2 region of LiTer4 are not distributed regularly in that region (Figure 3A). We took advantage of this irregular spacing to test the capacity of a segment carrying 3 matS sites to prevent MukBEF activity and to explore whether the density of matS sites can modulate MukBEF activity. To address these two questions, we deleted either matS26 or matS28from strain LiTer4 and probed using HiC the long-range of contacts. In strain LiTer4 ΔmatS26, three matS sites are distant from 78 and 58 kb whereas the matS sites in LiTer4 ΔmatS28strain are separated by 13 and 65 Kb. The ratio of normalized contact maps of the LiTer4 ΔmatS26 or LiTer4 ΔmatS28 strains to the corresponding ΔmatP derivative (Figure 3B, 3C and figure supplement 4C) revealed that long-range contacts were affected by the presence of only 3 matS sites in the 136-kb segment. By measuring the range of contacts in strains LiTer4 ΔmatS26 and LiTer4 ΔmatS28 (Figure 3E), the density of three matS sites ( matS25-matS26 and matS27) in strain LiTer4 ΔmatS28 seems to be as efficient (range contact of 357 kb +/- 11 kb) as the four matS sites in strain LiTer4 (range contact of 352 +/- 12kb). By contrast, in LiTer4 ΔmatS26, the range of contact is increased to 370 kb +/- 6 kb indicating a reduced ability to prevent MukBEF activity. Altogether, assuming that differences in the matS sequences do not modify MatP’s ability to bind to the chromosome and affect its capacity to inhibit MukBEF, these results suggested that the density of matS sites in a small chromosomal region has a greater impact than dispersion of the same number of matS sites over a larger segment.

MukBEF preferentially binds in newly replicated regions

Results presented above indicated that MukBEF activity can be initiated from multiple regions of the chromosome and that, unlike Smc-ScpAB, MukBEF does not initiate its activity in the Ori region and translocate linearly to the terminus of the chromosome. To further characterize the loading and translocation process of MukBEF, we performed ChIP-seq experiments using a Flag version of MukB on synchronized cells using a dnaC2 thermosensitive allele that allows to control the timing of replication initiation (Figure 4A). The cells were grown at a permissive temperature, then shifted to 40°C for 2 hours to allow the ongoing round of replication to complete without being able to initiate a new round; the cells were then shifted back to 30°C and samples were taken for ChIP-seq analyses after 10, 20 and 40 minutes. At t0, there is no variation in the number of sequences along the chromosome indicating the absence of replication. In this condition, the ChIP-seq signals show a slight enrichment of signals outside Ter. After 10 minutes at the permissive temperature, a fraction of >500 kb centered on oriC was replicated as revealed by an increase of sequencing reads in this region. After 20 minutes, a large zone of over 1.4 Mb was replicated, and after 40 minutes, the chromosome was fully replicated and the replication profile was similar to that of non-synchronized cells because of multiple new replication cycles initiated. Normalized ChIP-seq experiments were performed by normalizing the quantity of immuno-precipitated fragments to the input of MukB-Flag. This experiment showed a 2- to 4-fold enrichment in the regions that have been replicated (Figure 4A). After 10 min, the signal was increased over 500 kb on each side of oriC. At 20 min, the signal progressed and corresponded to the regions that have been replicated. At 40 min, when the chromosome has been fully replicated, the signals obtained in the ChIP-seq samples indicated an enrichment of MukBEF all along the chromosome except in a 1.5 Mb region centered on the Ter region. As shown in Figure 4B, MukBEF enrichment drops to basal levels 250kb before Ter (Figure 4A). This enrichment followed the progression of replication and spread from oriC towards Ter.

MukBEF preferentially binds in newly replicated regions and is exclude from Ter sequence

(A) ChIP analysis of mukB-Flag dnaC2 strain. Exponentially growing cells were synchronized by incubating them at 40°C for 2 hours, followed by a shift to 30°C for 0, 10, 20, or 40 minutes. The replication progression was monitored by plotting the input sequence (left panel) in 50 kb bins. Normalized ChIP (ChIP over input) values for 50 kb bins are presented in the right panel, with the red asterisk indicating the peak observed in all E. coli ChIP-seq experiments. The positions of oriC and Ter are highlighted with a red dashed line and on the chromosome schematic below the figure. To rule out the hypothesis of non-specific antibody binding dependent on replication, we conducted a ChIP experiment without tags in synchronized cells and did not detect enrichment comparable to what is observed here (figure supplement 9).

(B) ChIP analysis was conducted on the LiTer7 mukB-Flag dnaC2 strain. We observed that replication restart in this transposed strain exhibited a 20-minute lag. Therefore, cells in replication stop state were shifted to 30°C for 60 minutes to achieve a comparable replication progression to the 40-minute WT strain, as shown by the plotting of the input sequence (left panel). MukB enrichment generally followed the replication progression, except in the two Ter regions, as shown in the MukB normalized ChIP (right panel).

(C) Normalized Hi-C 5 kb bin contact map of ΔmukF pPSV38::mukBEF dnaC2 strain. Exponential cells were incubated at 40°C for 2 hours to prevent replication initiation (non-replicating condition) and were then shifted to 30°C for 40 minutes (right panel). MukBEF induction was performed by adding IPTG to the media for 2 hours. Quantification of long-range DNA contacts is shown in figure Supplement 8. Red dashed lines indicate the oriC position.

Altogether, these findings indicate that MukBEF is loaded into regions newly replicated either at the replication fork or even further behind it, except in the Ter region from which it would be excluded.

Ter segments prevent MukBEF binding

To further explore the ability of MatP to exclude MukBEF from the Ter region, we tested the ability of MatP to exclude MukBEF from chromosomal regions containing matS sequences. Chip-Seq experiments were performed on the Liter7 strain, which has a Ter2 segment closer to oriC. As in the wild-type background, MukBEF is preferentially associated with newly replicated sequences and shows a 2-fold increase following replication fork progression. However, a break in this enrichment profile is detected in the sequence corresponding to Ter2 (Figure 4B). These results suggest that MukBEF does not bind or persist in segments carrying matS sites, and that the MatP/matS system prevent residence of MukBEF in that region.

MukBEF promotes long-range contacts in the absence of replication

Since MukBEF was shown to bind preferentially in newly replicated regions, we wanted to test if a preferential activity of MukBEF was detectable in those regions. To do this, we performed Hi-C on non-replicating cells lacking MukF, and induced mukBEF expression with or without restarting replication. In the absence of replication and of MukBEF, long-range contacts were constrained by barriers that delimit domains, called Chromosome-Interacting Domains (CIDs), previously detected in WT cells in growing conditions (Lioy et al., 2018; figure supplement 5). Remarkably, after two hours of MukBEF induction in non-replicating cells, long-range contacts were detected except in the Ter region, indicating that MukBEF activity does not require newly replicated DNA to promote long-range contacts (Figure 4C, Figure Supplement 8). Finally, restarting replication for 40 minutes after 2 hours of MukBEF induction did not alter significantly the range and distribution of long-range DNA contacts observed in the absence of replication but in the presence of MukBEF outside Ter. As previously observed by microscopy in E. coli, replication does not appear to be an essential process for the activity of this SMC complex (Badrinarayanan et al., 2012a). Altogether, these results indicate that MukBEF promotes long-range DNA contacts independently of the replication process even though it binds preferentially in newly replicated regions.

Functional implications, comparative genomics of matS distribution

In E. coli, there are 28 matS sequences dispersed throughout the 1.03 Mb Ter domain. The average matS density is 3 matS per 100 Kb, but this distribution is not uniform and the matS density doubles in the vicinity of the dif sequence (figure supplement 6A). Despite the fact that only 9 matS are sufficient to completely inhibit MukBEF activity, matS sites have been selected and consequently a large portion of the E. coli chromosome is inaccessible to MukBEF.

To determine if other species also possess a large Ter domain, we used the matS consensus sequence to identify the Ter domain in 16 bacteria from the enterobacterales, pasteurellales, and vibrionales groups with the higher number of sequenced genomes. The Ter domain was defined as the longest stretch of matS sequences between areas that are at least 100 kb devoid of matS. The size of the Ter domain varies from 300 kb to 1 Mb, representing 6 to 25% of the chromosome, and contains 6 to 77 matS (Table 1). In most of these species, more than 14 matS sequences are distributed over 540 kb, and the Ter domain always contains the dif sequence. To test if the matS distribution might differ in the vicinity of the dif sequence, we measured the matS density and centered the distribution on the dif sequence. As shown in Figure 5, the number of matS per kb increases for all species and reaches its maximum near the dif site.

Distribution of matS sites culminates at dif across γ-proteobacteria

The density of matS sequences over a 100 kb-unit region was measured in 16 γ-proteobacteria, and the resulting distribution is plotted in the figure. The red line represents the 75th percentile of this distribution, while the blue line represents the 25th percentile. The light green area between these two lines represents the 50% of values closest to the median. The distribution is centered on the dif site, and the x-axis represents the genomic distance in megabases (Mb) from dif, while the y-axis represents the number of matS sequences per 100 kilobases (kb).

Discussion

SMC complexes play a fundamental role in the organization of genomes in all domains of life. Different models involving their loading, translocation or extrusion of DNA loops as well as their unloading have been proposed, based on results obtained with different complexes and in different models. Among the SMC complexes, MukBEF has specific features: MukBEF is only found in enterobacteria and some related bacterial genera where it is involved in chromosome segregation; the mukBEF genes belong to a group of genes co-occurring with the Dam methylase gene including also matP and other genes involved in DNA metabolism (Brezellec et al., 2006); MukBEF exists as dimers of dimers connected by the kleisin subunit MukF (Badrinarayanan et al., 2012b; Zawadzka et al., 2018); by its activity, MukBEF does not align the two arms of the chromosome like the canonical bacterial Smc-ScpAB complex but instead promotes long distance contacts in cis like a eukaryotic condensin (Lioy et al., 2018). Remarkably, MukBEF activity is not detected within the specific chromosomal Ter domain, one-fifth of the E. coli chromosome, due to the presence of MatP associated with this region. Although details about MukBEF and its activity have been revealed in recent years, key steps in its function remain to be characterized, including loading onto the DNA molecule, its actual loop extrusion activity, and its unloading by MatP.

Altogether, our data provide an integrated view of MukBEF activity to organize the E. coli chromosome. By different ChIP-seq and HiC approaches performed in different strains, some of which have undergone programmed genetic rearrangements, we showed that MukBEF loading does not only involve the Ori region but also different regions of the chromosome. Our results also indicated that although MukBEF binds preferentially in newly replicated regions, its activity is detected even in the absence of replication and long-range contacts appear similarly throughout the genome, except in the Ter region. These results support a model in which MukBEF molecules are bound to the chromosome, molecules are removed or displaced by replication, MukBEF molecules readily reassociate in newly replicated regions except in Ter region in which the unloading of MukBEF is enhanced by MatP bound to matS sites. Our results are in agreement with the previous proposal that MukBEF may organize a series of loops around a thin MukBEF axial core (Mäkelä and Sherratt, 2020). Altogether, our results reveal a striking contrast with the way Smc-ScpAB loads on DNA by interacting with the ParB at parS sites and then translocates towards the ter region (Wang et al., 2017). Instead, long-distance contacts promoted by MukBEF did not occur as a wave from a specific region but rather initiate at different positions, in different regions of the genome. Thus, the activity of MukBEF appears to be more similar to that of eukaryotic condensins than to that of the Smc-ScpAB complex.

The fact that MukBEF is found associated with newly replicated DNA could also suggest preferential loading at the replication fork or behind. However, in the absence of replication, MukBEF is still able to load and form long-range DNA contacts, as well as segregated chromosomes (Badrinarayanan et al., 2012a). Interestingly, this situation is reminiscent to that of SMC-ScpAB. Smc-ScpAB is loaded onto the B. subtilis chromosome at the parS site in association with parB/SpoJ (Gruber and Errington, 2009; Sullivan et al., 2009). Remarkably, in the absence of parB, Smc-ScpAB is still able to perform chromosome segregation almost perfectly (Ireton et al., 1994; Wang et al., 2014). This demonstrates that, even with a strong preferential loading mechanism, the SMC complex can load on chromosomal DNA. A similar situation may occur in E. coli, where MukBEF is preferentially loaded depending on the progression of replication but can still bind to and act on the chromosome in the absence of replication.

Our results showed that MukBEF was not detected in regions containing matS sites and that these regions were devoid of contacts extending over 600 kb. By varying the number of matS sites at an ectopic position, we showed that MukBEF inhibition is graded with the number of matS; while an effect is already visible with 3 matS scattered over a 78 kb region, the maximal effect seems to be reached with 9 matS sites distributed over a 253 kb region. As proposed for Smc-ScpAB and its unloading XDS site (Karaboja et al., 2021), if the movement of DNA in the loop extrusion process involves large steps, it is conceivable that several matS sites are required for a complex to be trapped and discharged by MatP bound to a matS site. Accordingly, the density of matS sites seems to affect the inhibition efficiency supporting this assumption. Further experiments will be needed to analyze in details the optimal spacing of matS sites to inhibit MukBEF activity.

The structure of the MukBEF-AcpP-MatP/matS complex obtained by cryoEM revealed the entrapment of two topologically separated DNA segments in two distinct compartments called “ring” and “clamp”, with the matS site present in the ring compartment. The proposed model for MukBEF unloading stipulates that the unloading of the segment carrying matS site in the ring compartment is coupled to the unloading of the other segment in the clamp. One of the MatP monomers forms a contact with one of the MukE monomers while the joint binds and positions MatP between the MukB arms. The joint interface is much larger than the MukE-MatP bridge and likely provides the major binding energy for association (Bürmann et al., 2021). Indeed, a change to alanine of the five residues between H38 and D42 in contact with MukE did not affect MatP’s ability to inhibit MukBEF activity (figure supplement 7). Refined experiments will be required to assess the outcome of mutating MatP residues involved in the interaction with the joint as mutations of those residues also affect matS binding (Dupaigne et al., 2012).

MukBEF activity is detected by the appearance of contacts over a distance of approximately one megabase, spanning around 4 megabases of the chromosome (excluding the Ter domain). Even if the molecular details of this activity remain to be characterized, it is tempting to speculate that MukBEF molecules can be discharged from DNA in the absence of MatP, i.e. outside Ter. Furthermore, it can be noted that MukBEF activity does not appear to be significantly disrupted in a matP mutant suggesting that MukBEF can be discharged from DNA in the absence of MatP. As proposed by Bürmann and colleagues (Bürmann et al., 2021), MatP would act as a structural element that ensures ideal positioning of DNA close to the exit gate in the MukBEF complex that might ensure an efficient unloading of MukBEF from the DNA. Altogether, the results would indicate that MatP is not the MukBEF unloader per se, but rather that its ability to prevent MukBEF activity has been selected to protect the Ter region from a condensin activity.

The distribution of matS sites over a large region of the chromosome in enterobacteria results in the inhibition of MukBEF in Ter. MatP has already been shown to confer another property to Ter, through an interaction of its C-terminus with the septum-associated protein ZapB and localizing Ter at midcell. Remarkably, these two properties are independent as inhibiting MukBEF does not require its anchoring of Ter at the septum of division (Lioy et al., 2018). Given that the number of matS sites far exceeds the number required to prevent MukBEF activity and that the density of matS sites increases as we approach dif, we may speculate that MukBEF presence is mostly banished from the dif region. At least two activities of DNA metabolism occur in this region: resolution of chromosome dimers by XerC-XerD recombinases and the post-replicative decatenation of circular chromosomes. We may speculate, based on in vitro observations (Kumar et al., 2022), that MukBEF could interfere with TopIV activity and delay potential chromosome decatenation. Another possibility is that chromosome dimers resolved at the dif site may become trapped in loops formed by MukBEF, thus delaying segregation. However, none of these possible scenarios are supported by data yet, and a major challenge for the future will be to determine whether and how MukBEF may interfere with one or both of these processes.

In the absence of replication and of MukBEF activity, the contact map displayed a single strong diagonal composed of very well-defined CIDs (for Chromosomal Interacting Domains), ranging in size from 20 to 400 kb. The CIDs, clearly visible in populations of cells growing exponentially, were prominent in these conditions revealing the impact of transcription on the structuring of the genome. Upon replication, the CID organization is less apparent because of the new contacts that occur between loci belonging to different CIDs. The induction of MukBEF in the presence of replication further attenuates the patterning of CIDs. Altogether, these results highlight the respective contribution of these three processes, transcription, replication and condensin activity, on the organization of bacterial genomes.

Materials and methods

Bacterial strains and plasmids

All E. coli strains used in this study are derived from E. coli MG1655. Deletion mutants were constructed as described in (Datsenko and Wanner, 2000). Mutations were combined via P1 transduction. A plasmid capable of synthesizing MukBEF was constructed by cloning the entire mukBEF operon into the pPSV38 plasmid using EcoRI/XbaI. MukBEF synthesis was monitored by western blot after the addition of IPTG at 0, 20, 40, and 120 minutes (Figure Supplement 1D).

Media and growth conditions

E. coli cells were cultured at 22°C and 30°C in either Lennox Broth (LB) or liquid minimal medium A (MM) supplemented with 0.12% casamino acids and 0.4% glucose. When necessary, antibiotics were added to the growth media at the following concentrations: ampicillin (Amp) at 100 μg/mL, kanamycin (Kan) at 50 μg/mL, chloramphenicol (Cm) at 15 μg/mL, spectinomycin (Sp) at 50 μg/mL, apramycin (Apra) at 50 μg/mL, and zeocin (Zeo) at 25 μg/mL.

3C-seq protocol

3C-seq librairies were generated as described (Lioy et al., 2018). Briefly 100ml of culture were crosslinked with formaldehyde (7% final concentration) for 30 min at room temperature (RT) followed by 30 min at 4°C. Formaldehyde was quenched with a final concentration of 0.25 M glycine for 20 min at 4°C. Fixed cells were collected by centrifugation and stored at −80°C until use. Frozen pellets consisting of approximately 1-2 x 109 cells were thawed and suspended in 600 μl of TE buffer (10 mM Tris-HCl, 0.5 mM EDTA, pH 8) with 4 μl of lyzozyme (35 U/μl). The mixture was incubated at room temperature for 20 minutes. Subsequently, SDS was added to a final concentration of 0.5% and the cells were incubated for an additional 10 minutes at room temperature. Lysed cells were then diluted 10 time in several tubes containing 450 μl of digestion mix (1X NEB 1 buffer, 1% Triton X-100). 100 units of HpaII were added and the tubes were incubated for 2 hours at 37°C. To stop the digestion reaction, the mixture was centrifuged for 20 minutes at 20,000 g, and the resulting pellets were resuspended in 500 μl of sterile water. The resulting digested DNA (4 ml in total) was divided into four aliquots and diluted in 8 ml of ligation buffer (1X ligation buffer NEB without ATP, 1 mM ATP, 0.1 mg/mL BSA, 125 units of T4 DNA ligase, 5 U/μl). Ligation was performed at 16°C for 4 hours, followed by overnight incubation at 65°C with 100 μl of proteinase K (20 mg/ml) and 100 μl EDTA 500 mM. DNA was then precipitated with an equal volume of 3 M Na-Acetate (pH 5.2) and two volumes of iso-propanol. After 1 hour at −80°C, the DNA was pelleted and suspended in 500 μl of 1X TE buffer. The tubes were incubated directly with 50 μl of proteinase K (20 mg/mL) for an overnight period at 65°C. Subsequently, all tubes were transferred to 2 ml centrifuge tubes and extracted with 400 μl of phenol-chloroform pH 8.0. The DNA was then precipitated, washed with 1 ml of 70% cold ethanol, and resuspend in 30 μl of 1X TE buffer in the presence of RNase A (1 μg/ml). The tubes containing ligated DNA (3C libraries), digested DNA, and non-digested DNA were pooled into three separate tubes. The efficiency of the 3C preparation was evaluated by running a 1% agarose gel.

Hi-C protocol

Hi-C librairies were generated as described (Thierry and Cockram, 2022).

108 cells growing in the exponential growth phase of E. coli were chemically crosslinked by the addition of formaldehyde directly to the cultures (3% final concentration) for 30 min at room temperature with gentle agitation. Crosslinking was quenched by the addition of glycine (0.5 M final concentration) for 20 min at room temperature. Celle were wash in 50 ml PBS 1X and centrifuged. The pellet was resuspended in 1 ml 1 x PBS and transferred to a 1.5 ml Eppendorf tube before a final centrifugation step (4000 x g, 5 min, room temperature), the supernatant was then removed and the pellet stored at −80°C. The pellet was then resuspended in 1 ml of 1x TE + complete protease inhibitor cocktail (EDTA-free, Sigma Aldrich) and transferred to a 2 ml Eppendorf tube with 4 µl of Ready to lyse Lysozyme for 20mn at room temperature. SDS was added to a final concentration of 0.5% and the cells were incubated for an additional 10 minutes. In a 10 ml Falcon tube, DNA was then prepared for digestion by the addition of 3 ml H2O, 500 μl 10X Digestion buffer (200 mM Tris-HCl pH 7.5, 100 mM MgCl2, 10 mM DTT, 1 mg/ml BSA) and 500 μl 10% Triton X-100 (Thermo-Fisher). After thoroughly mixing the reaction, 400 μl were removed and transferred to a 1.5 ml Eppendorf tube as a non-digested (ND) control. The restriction enzyme HpaII (New England Biolabs, 1000 U) was then added to the remaining sample and the tube incubated with gentle agitation for 3h at 37°C. The solution was centrifuged (16,000 x g, 20 min, room temperature. After removing the supernatant, the pellet was resuspended in 400 μl H2O, completed subsequently by adding 50 μl 10x Ligation Buffer (500 mM Tris-HCl pH 7.5, 100 mM MgCl2, 100 mM DTT), 4.5 μl 10 mM dAGTTp, 37.5 μl Biotin-14-dCTP (Thermo Fisher), 50 Units of DNA Polymerase I - Large Klenow Fragment (New England Biolabs). After briefly mixing, the reaction was incubated with agitation for 45 min at 37°C. The ligation was set up by adding the following; 120μl 10x Ligation Buffer, 12 μl 10 mg/ml BSA, 12 μl 100mM ATP, 540 μl H2O, 480 U T4 DNA ligase (Thermo Fisher). The reaction was mixed gently and then incubated with gentle agitation for 3h at room temperature. Following ligation, proteins were denatured by the addition of 20 μl 500 mM EDTA, 20 μl 10% SDS, and 100 μl 20 mg/ml proteinase K (EuroBio). The following day, DNA was purified using the standard procedure describe for the 3C-seq.

Chip-seq protocol

109 cells in the exponential growth phase of E. coli were chemically crosslinked by adding formaldehyde directly to the cultures (final concentration of 1%) for 30 minutes at room temperature with gentle agitation. Crosslinking was quenched by adding glycine (final concentration of 0.25 M) for 15 minutes at room temperature. The cells were washed twice with 1X TBS (50mM Tris-Hcl ph 7.6, 0.15M NaCl) and the pellet was resuspended in 1 mL of 1X TBS before a final centrifugation step (4000 x g, 5 minutes, room temperature). The supernatant was then removed and the pellet was stored at −80°C.

The pellet was resuspended in 500 µl of lysis buffer 1 (20% sucrose, 10 mM Tris pH 8, 50 mM NaCl, 10 mM EDTA) and 4 µl of ready-to-lyse lysozyme was added, followed by incubation at 37°C for 30 minutes. Then, 500 µl of lysis buffer 2 (50 mM Tris-HCl pH 4.7, 150 mM NaCl, 1 mM EDTA, 1% Triton X100) and a tablet of antiprotease cocktail (Roche) were added. The solution was transferred to a 1 mL Covaris tube and sonicated for 10 minutes (peak incident power 140 W/duty cycle 5%/cycle per burst 200). Cell debris were eliminated by centrifugation at 13,000 rpm for 30 minutes at 4°C. The supernatant was transferred to a fresh Eppendorf tube and 50 µl was used as input. The rest of the cell extract was mixed with 40 µl of anti-Flag M2 resin previously washed in TBS and resuspended in lysis buffer 2. The solution was incubated overnight at 4°C with gentle mixing and then centrifuged for 30 seconds at 5,000 g. The pellet was washed twice with TBS containing 0.05% Tween 20 and three times with TBS.

To elute the immunoprecipitation, 100 µl of 1X TBS containing 15 µg of 3X Flag peptide was mixed with the resin and incubated for 30 minutes at 4°C. Then, the solution was centrifuged for 30 seconds at 5,000 g and the supernatant was transferred to a new tube. A second elution step was performed on the resin before decrosslinking. IP and input was purified using MinElute Qiagen columns and then sequenced.

Processing of libraries for Illumina sequencing

The samples were sonicated using a Covaris S220 instrument to obtain fragments ranging in size from 300 to 500 base pairs. These fragments were then purified using AMPure XP beads and resuspended in 10 mM Tris-HCl. For Hi-C libraries, a biotinylated pull-down step was performed by adding 30 µl of streptavidin C1 MyOne Dynabeads from Invitrogen to 300 µl of binding buffer (10 mM Tris-HCl, pH 7.5, 1 mM EDTA, 2 M NaCl) and mixing with the Hi-C libraries for 15 minutes. DNA ends were then prepared for adaptor ligation following standard protocols as described in (Thierry and Cockram, 2022).

The Illumina sequencing process was performed in accordance with the manufacturer’s recommendations, with 15 cycles of amplification. The size of the DNA fragments in the libraries was assessed using TAE 1% agarose gel and tape station, followed by paired-end sequencing on an Illumina sequencer.

Chip-seq analysis

Between 10 to 20 million reads were recovered for each sample. We used Bowtie2 software to perform mapping in local mode, and the mpileup software from Samtools to calculate coverage for each position in the genome. Normalized ChIP was then calculated by normalizing the number of reads at each position by the total number of reads, and dividing this number by the normalized input. A sliding window of 50 kb was applied to smooth the variations. Peaks for MatP ChIP-seq (figure supplement 3B) were identified by extracting positions where the number of reads was ten times higher than the background for at least 30 consecutive base pairs. The center of these distributions was considered the center of the peak. The sequences were then extracted and used on the MEME suite to identify a common motif.

Generation of contact maps

Contact maps were constructed as described previously (Lioy et al., 2018). Briefly, each read was assigned to a restriction fragment. Non-informative events such as self-circularized restriction fragments, or uncut co-linear restriction fragments were discarded, as in (Cournac et al., 2012). The genome was then divided into 5 kb units, and the corresponding contact map was generated and normalized through the sequential component normalization procedure of SCN (based on the sequential component normalization; https://github.com/koszullab/E_coli_analysis) (Lioy et al., 2018). Contact maps were visualized as logarithmic matrices to aid in visualization.

Ratio of contact maps

The comparison of contact maps was facilitated by displaying their ratio. The ratio was calculated for each point on the map by dividing the number of contacts in one condition by the number of contacts in the other condition. The Log2 of the ratio was then plotted using a Gaussian filter. The color code represented a decrease or increase in contacts in one condition relative to the other (a blue or red signal, respectively); a white signal indicated no change.

Quantification of the range of cis contacts along the chromosome

We used a three-step process adapted from (Lioy et al., 2020) and (Wang et al., 2017) to determine the width of the diagonal in contact maps. To improve the resolution of the Ter limit, we measured the width perpendicularly to the main diagonal.

First, we calculated the median of the contact map and estimated the standard deviation (σ) using a robust statistic, where σ = 1.4826 * mad (mad = median absolute deviation). Next, we used a point connecting algorithm to differentiate significant interactions from background noise. The size of each connected element identified by the “bwlabel” function of MATLAB was determined. We considered a connected element with a size greater than or equal to 30 points to be significant and used the “imclose()” function of MATLAB to fill in the empty points within the connected elements using a diamond shape with a size of 5.

Subsequently, we calculated the width of the primary diagonal for each 5-kb bin of the genome. The range of cis contact was estimated from the width of the primary diagonal by multiplying the number of measured bins by the bin size (5 kb) and dividing by two (since the range is symmetric on both sides of the chromosomal locus being considered). Finally, a boxplot was used to visualize the entire range of cis contacts for all the analyzed chromosome regions.

Directional Index analysis

Directional Index was calculate as described in (Lioy et al., 2018). The directional index is a statistical metric that quantifies the level of upstream or downstream contact bias for a given genomic region (Dixon et al., 2012). This metric is based on a t-test comparison of contact vectors to the left and right of each bin, up to a certain scale. The boundaries between topological domains often generate fluctuating signals that result in a transition in the directional preference. Specifically, for each 5 kb bin, we extracted the contact vector from the correlation matrix between that bin and neighboring bins at regular 5 kb intervals, up to 100 kb, in both left and right directions. At each step, the paired t- test was used to determine whether the strength of interactions was significantly stronger in one direction relative to the other. A threshold of 0.05 was used to assess statistical significance. The directional preferences for each bin along the chromosome were visualized as a bar plot, with positive and negative t-values shown as red and green bars, respectively. To improve the clarity of presentation, bars for bins with t-values below −2 or above 2 (corresponding to a p-value of 0.05) were truncated. Between two identified domains in the contact matrices, the directional preference of bins changed, which was indicated by alternating red and green colors.

Data availability

The DNA sequencing raw data was deposited on the NCBI website under the bioproject numbers: PRJNA1019269 (HiC induction of MukBEF), PRJNA1021387 (HiC on transposed strains), PRJNA1024671 (HiC on non-replicating strains), and PRJNA1024669 (ChIP-seq of MukB on synchronized strains).